### 2012-03-24 comparison.tex
updated lineup section

\documentclass[journal]{vgtc}                % final (journal style)
%\documentclass[review,journal]{vgtc}         % review (journal style)
%\documentclass[widereview]{vgtc}             % wide-spaced review
%\documentclass[preprint,journal]{vgtc}       % preprint (journal style)
%\documentclass[electronic,journal]{vgtc}     % electronic version, journal

%% Uncomment one of the lines above depending on where your paper is
%% in the conference process. ``review'' and ``widereview'' are for review
%% submission, ``preprint'' is for pre-publication, and the final version
%% doesn't use a specific qualifier. Further, ``electronic'' includes
%% hyperreferences for more convenient online viewing.

%% Please use one of the ``review'' options in combination with the
%% assigned online id (see below) ONLY if your paper uses a double blind
%% review process. Some conferences, like IEEE Vis and InfoVis, have NOT
%% in the past.

%% Please note that the use of figures other than the optional teaser is not permitted on the first page
%% of the journal version.  Figures should begin on the second page and be
%% in CMYK or Grey scale format, otherwise, colour shifting may occur
%% during the printing process.  Papers submitted with figures other than the optional teaser on the
%% first page will be refused.

%% These three lines bring in essential packages: ``mathptmx'' for Type 1
%% typefaces, ``graphicx'' for inclusion of EPS figures. and ``times''
%% for proper handling of the times font family.

\usepackage{natbib}
\usepackage{mathptmx}
\usepackage{graphicx}
\usepackage{times}
\usepackage{amsmath}

%% We encourage the use of mathptmx for consistent usage of times font
%% throughout the proceedings. However, if you encounter conflicts
%% with other math-related packages, you may want to disable it.

%% If you are submitting a paper to a conference for review with a double
%% blind reviewing process, please replace the value ``0'' below with your
%% OnlineID. Otherwise, you may safely leave it at ``0''.
\onlineid{0}

%% declare the category of your paper, only shown in review mode
\vgtccategory{Research}

%% allow for this line if you want the electronic option to work properly
\vgtcinsertpkg

%% In preprint mode you may define your own headline.
%\preprinttext{To appear in an IEEE VGTC sponsored conference.}

%% Paper title.

\title{Graphical Tests for Power Comparison of Competing Designs}

%% This is how authors are specified in the journal style

%% indicate IEEE Member or Student Member in form indicated below
\author{Lendie Follett, Heike Hofmann, Mahbub Majumder, Dianne Cook}
\authorfooter{
%% insert punctuation at end of each item
\item
 Lendie Follett is a graduate student in Statistics at Iowa State University, E-mail: lfollett@iastate.edu.
\item
everybody else is at Iowa State University, too
}

%other entries to be set up for journal
\shortauthortitle{Follett \MakeLowercase{\textit{et al.}}: Comparison of Competing Layouts}
%\shortauthortitle{Firstauthor \MakeLowercase{\textit{et al.}}: Paper Title}

%% Abstract section.
\abstract{
Lineups  \cite{buja:2009, wickham:2010} have been established as tools for visual testing
similar to standard statistical inference tests, allowing us to
evaluate the validity of graphical findings in an objective manner. In
simulation studies  \cite{majumder:2011} lineups have been shown as being efficient: the
power of visual tests is comparable to classical tests while being
much less stringent in terms of distributional assumptions made. This
makes lineups versatile, yet powerful, tools in situations where
conditions for regular statistical tests are not or cannot be met. In
this paper we introduce lineups as a tool for evaluating the power of
competing graphical designs. We highlight some of the theoretical
properties and then show results from two studies evaluating competing
designs: both studies are designed to go to the threshold of our
perceptual abilities to highlight differences between designs. We use
both accuracy and speed of evaluation as measures of a successful
design. The first study compares the choice of coordinate system:
polar versus Euclidean coordinates. The results show strong support in
favor of Euclidean coordinates in finding fast and accurate answers to
spotting patterns. The second study is aimed at finding shift
differences between distributions. Both studies are motivated by data
problems that we have recently encountered, and explore using
simulated data to evaluate the plot designs under controlled
conditions.  Amazon Turk is used to conduct the studies.  The lineups
provide an effective mechanism for objectively evaluating plot
designs.
} % end of abstract

%% Keywords that describe your work. Will show as 'Index Terms' in journal
%% please capitalize first letter and insert punctuation after last keyword
\keywords{Lineups, visual inference, Power Comparison, Efficiency of Displays.}

%% ACM Computing Classification System (CCS). 
%% See <http://www.acm.org/class/1998/> for details.
%% The ``\CCScat'' command takes four arguments.

\CCScatlist{ % not used in journal version
  \category{H.5.2}{Information Interfaces and Presentation}{ User Interfaces --- Graphical user interfaces (GUI), Interaction styles, Screen design, Evaluation/methodology}
  \CCScat{I.6.8}{Computing Methodologies}%
  {Simulation and Modeling}{Visual Simulation};
}

\graphicspath{{images/}}

%% Uncomment below to include a teaser figure.
\teaser{
\centering
\includegraphics[height=1in]{polar}\hspace{0.05\linewidth}
\includegraphics[height=1in]{eucline}
%\includegraphics[width=16cm]{CypressView.eps}
\caption{teaser chart should include all different designs}
}

%% Uncomment below to disable the manuscript note
%\renewcommand{\manuscriptnotetxt}{}

%% Copyright space is enabled by default as required by guidelines.
%% It is disabled by the 'review' option or via the following command:
% \nocopyrightspace

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%% START OF THE PAPER %%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%% The ``\maketitle'' command must be the first command after the
%% ``\begin{document}'' command. It prepares and prints the title block.

%% the only exception to this rule is the \firstsection command
\firstsection{Introduction}

\maketitle

%% \section{Introduction} %for journal use above \firstsection{..} instead
%We introduce and discuss a tool for comparing different charts that are renderings of the same data in order to evaluate them for efficiency.
Graphics have a central role in the discovery process as well as in the presentation of results. 
As designer of  charts one of our main goals is to display data so that the information it contains is relayed in the most efficient way. 
In the search for the best display out of the seemingly endless multitude of methods we are guided by experience, knowledge of perceptual strengths and weaknesses, as well as aesthetic preferences. It is hard, if not impossible, to make the final decision on the design in an  objective manner. 
We can make this process less subjective by conducting user studies to evaluate competing designs of the same data, and measure designs for their efficiency by recordung how accurately and quickly viewers can extract pieces of information. 
User studies for evaluating different designs has a long standing tradition. For  statistical graphics user studies have their origin in the (convenience) study on evaluating difficulty of a set of different visual tasks by \citet{cleveland:1984}, which has recently been repeated finding similar results based on a larger and more representative study by \citet{kosara:2010}.

Usually, we need simulation studies to be able to completely control for the signal strength in the data. 
For charts, this is not a very realistic scenario. The need for displaying information particularly efficiently or accurately usually arises from  a particular problem while analyzing a data set, from which we want to learn new insight. Often, it is difficult to simulate data in such a way that it realistically represents the particular problem, which in turn makes it dubious whether the results from a simulation study are actually applicable to the problem at hand, reminding us painfully of George Box's famous quote on models: ``essentially, all models are wrong, but some are useful." \citep[page=424]{box}

Lineup tests  \citep{buja:2009} allow us to directly use the data at hand to evaluate different competing designs by using permutations tests \cite{good:2011} of the original dataset. We can therefore get valid quantitative measures for comparing different designs without having to set up a simulation.

When we are presenting a chart we  want to convince our audience of the presence of a particular feature or relationship in the data. 
In order to enable us to draw a conclusion about presence or absence of a feature in the data, we have, however, to assess what features show up based purely on marginal distributions rather than joint distributions.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[height=0.32\linewidth]{rexp.pdf} 
   \includegraphics[height=0.32\linewidth]{rnorm.pdf} 
   \includegraphics[height=0.32\linewidth]{two-exp.pdf} 
   \caption{Examples of charts with distinctive features but independent variables. From left to right: scatterplot of $X$ and $Y$ with independent $X, Y \sim Exp_{\lambda=1}$; scatterplot of two variables with mixed normal distribution $N(3,1)$ and $N(-3,1)$; boxplot of $X$ and $Y$ with $X \sim B_{n=100, p=0.1}$ and $Y \sim Exp_{\lambda=1}$. }
   \label{random}
\end{figure}

Features in the marginal distributions, such as multi-modality, skewness  or outliers, might interfere with our intuition of what `randomness' might mean in a particular situation and lead us to wrong conclusions. All of the plots in figure \ref{random} show distinctive features. None, of these, however, is a feature of the joint distribution, i.e. all of the variables involved are independent of each other. The scatterplot on the left of figure \ref{random} shows data sampled from an exponential distribution with rate parameter $\lambda=1$. The data in the middle comes from mixing two normal distributions with different means: half of the points has mean $\mu_1 = -3$, the other half has mean $\mu_2=3$, which results in the distinctive four groups.
The two boxplots on the right appear to show a median shift between the two groups. This, however is a result of both the group size (only 7\% of the points is in the second group) and a skew distribution in vertical direction, with $y \sim Exp_{\lambda=1}$.

Lineup plots have been shown as an effective tool \cite{wickham:2010} for enabling us to quantitively assess the strength of the relationship between $X$ and $Y$.
In a lineup plot the plot of the actual data is -- as in a police line-up -- inserted among decoy plots. The decoy or {\it null} plots are generated in a way that is consistent with the null hypothesis, i.e. these are plots in which we can assume that any relationship visible is there purely by chance. This gives us a reference framework against which to compare the data plot and, if we are still able to identify the data plot from the lineup because of a distinguishing feature, we have formal evidence that the feature in the data is, in fact, not just coincidental.

The strength of this evidence is determined by both the number of independent evaluations as well as the size of the lineup. (XXXX needs more motivation HH) The signal strength \cite{majumder:2011} of a particular lineup is given as:
\[
\text{signal strength} = 1- \left( \frac{x}{n}\right)^{1/(m-1)},
\]
where $m$ is the size of  the lineup, i.e. the number of individuals shown, $x$ is the number of times viewers identified the data plot correctly in  $n$ independent evaluations. Signal strength is limited to a range of values between 0 and 1. Higher values indicate stronger signal.
 In the case of a parametric situation (i.e. the feature shown in the chart can be described in parametric form as a real-valued vector $\theta$ in the data), the signal strength is equivalent to the statistical power of the test.

\section{Lineups for Comparing Designs}
We are going to make use of the signal strength gained from multiple viewings of a lineup in order to evaluate competing designs as follows:
\begin{enumerate}
\item{{\bf Create Lineup Data:} create $m-1$ permutations of $y$}
\item{{\bf Create lineups from competing designs:} use the same data}
\item{{\bf Evaluate Lineups:} assess both signal strength and time taken by individuals to come to a decision.}
\end{enumerate}
differences in signal strength are due to rendering (after correcting for individuals' visual ability)

XXXX should we include nullabor code?

%We also know that humans learn about new information through a visual search pattern that emphasizes new information (Wolfe, 2002; Rensick, et al., 1997). Of course, the type of visualization should take the natural human inclinations into account as well. For instance, a good visualization will account for both the ãgestalt? like perceptual tendencies as well as the detail-orientated nature of cognition (Kosslyn, Thompson and Ganis, 2002; Pylshyn,2003)

\section{Experimental Setup}
Describe the experimental setup: polar charts versus euclidean - with and without reference lines.
Figure showing all four types using the same dataset. 
Discuss construction of each one. Discuss construction of null chart.

The goal of this experiment is to understand the efficiency of displaying information using euclidian and polar coordinates. Efficiency is the deviation from independence, which can be simulated by taking permutations of the original dataset. Data concerning wind patterns and time between flights at SEA airport is used for these charts. The lineup chart is composed of 20 graphics with all of the exact same characteristics except for the data. One of the 20 graphics will be using original airport data while the other 19 will be permutations of that data. Efficiency of the chart types can be inferred by how accurately individuals pick out the plot with the original data. There are other variables which are introduced to further our understanding of the way we perceive euclidian and polar coordinates. All combinations of each level of the variables are involved in the experiment for both polar and euclidian coordinates. The graphics in each chart will all have exactly the same level of the variables with the only difference being the data.

The other variables which are monitored are existence of reference line, placement of x-axis, and sample size. It is possible that one of the chart types benefits more from additional components such as reference lines. The polar and euclidian null charts will be shown either with or without a white reference line. The line is placed at slightly over 50 percent of the y axis. Additionally, the x-axis has four placements. The original dataset contains a characteristic wave pattern, which for euclidian coordinates, may be a large factor contributing to the perceived effectiveness of the graphic. Characteristic patterns such as the wave pattern may be perceived in a different way when viewed in polar coordinates. When the x-axis is is shifted the look of the pattern changes. The wave pattern is broken up for euclidian coordinates but for polar coordinates is only shifted around the axis. Finally, the amount of information necessary for detection as well as correctness is a measure for efficiency of the charts so the sample size which was taken from the original dataset is of interest. The original data set contained XXX,XXX observations from which 2, 4, 6, 8, 10, and 24 percent of the observations were sampled. Small sample sizes are easier to obtain because of time and money constraints. It would be nice to know if there is one type of chart which can show data trends using less information.

 All combinations of these variables were repeated for plots with a reference line and without a reference line, as shown in figure \ref{layouts}.This figure shows examples of graphics showing the actual data, which in the experiment would be randomly positioned among 19 null plots. In total there are XXX unique charts which represent all different combinations of sample size, coordinate type, x-axis placement, and whether or not there is a reference line. The charts are all assigned a difficulty level, which is based on sample size, on a scale from zero to four. There are two plots assigned a difficulty level of zero. These two charts, one polar and one Euclidian, were made from computer generated data and intended to be at the level where an individual, if he or she is not purely guessing, should be able to find the correct plot.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=0.4\linewidth]{Euclidian_NoLine.pdf} \hspace{0.05\linewidth}
   \includegraphics[width=0.4\linewidth]{Euclidian_Line.pdf} \\
   \includegraphics[width=0.475\linewidth]{Polar_NoLine.pdf} 
   \includegraphics[width=0.475\linewidth]{Polar_Line.pdf}  
   \caption{All four layouts of charts The first row shows the data in polar coordinates and the second row shows the data in euclidian coordinates. The first column in both rows show the plots without any reference lines. The second column in both rows have a white reference line added. }
   \label{layouts}
\end{figure}

Amazon's Mechanical Turk is an online service in which individuals can request and perform simple tasks and surveys for money. Using Turk we were able to gather data on how people responded to the charts. Each person is shown a series of ten charts and asked to choose the graphic on each chart which they think is "different". They then identify, from a list of choices, why they chose this particular chart as well as how confident they are on a scale of one to five. Personal information such as age group, gender, and education is also collected on a voluntary basis. The amount of time it took the individual to answer is also recorded. This study allows us to make informed judgments about how the plot type effects efficiency in terms of accuracy and time spent reading the plots. Xcite stephen few?X

Effect is deviation from independence. Cannot be properly measured statistically except in highly aggregated situations, where the effect might be washed out.

Most generally this experiment can be paralleled to the pie chart versus barchart discussion. The Euclidian coordinate plot is made up of bars *whose?* areas are colored proportionally according to the data. The polar coordinate plot is made up of pie slices which are also colored proportionally according to the data. Viewers of these charts will visually decode the areas of the bars and pie slices and make a judgement based on their decoding. How accurate this judgment is is left to be discussed. According to (Kosslyn, Stephen, Graph Design for the Eye and Mind, Oxford University Press, 2006, p. 40), area is usually perceived as a function of what the actual area is. This function is the area raised to an exponent of approximately 0.8 and then multiplied by a constant. However, when bars are parallel, the relative height of these bars is perceived very accurately. From this information, one could infer that the Euclidian coordinate plot may gain efficiency based on the height of the colored bars.

Pie versus Barchart discussion is pretty old: cite some of the literature:

From Stephen Few, perceptual edge:
	When a graph is made, quantitative and categorical information is encoded by a display method. Then the information is visually decoded. This visual perception is a vital link. No matter how clever the choice of the information, and no matter how technologically impressive the encoding, a visualization fails if the decoding fails. Some display methods lead to efficient, accurate decoding, and others lead to inefficient, inaccurate decoding. It is only through scientific study of visual perception that informed judgments can be made about display methods. (William S. Cleveland, The Elements of Graphing Data, Hobart Press, 1994, p. 1)

	The systematic distortion of area is captured by ï¿½Stevenï¿½s Power Law,ï¿½ which states that the psychological impression is a function of the actual physical magnitude raised to an exponent (and multiplied by a scaling constant). To be precise, the perceived area is usually equal to the actual area raised to an exponent of about 0.8, times a scaling constant...In contrast, relative line length [such as the lengths of bars] is perceived almost perfectly, provided that the lines are oriented the same way. (Kosslyn, Stephen, Graph Design for the Eye and Mind, Oxford University Press, 2006, p. 40)

	We make angle judgments when we read a pie chart, but we donï¿½t judge angles very well. These judgments are biased; we underestimate acute angles (angles less than 90ï¿½) and overestimate obtuse angles (angles greater than 90ï¿½). Also, angles with horizontal bisectors (when the line dividing the angle in two is horizontal) appear larger than angles with vertical bisectors. (Naomi Robbins, Creating More Effective Graphs, Wiley, 2005, p. 49)

	 Edward Tufte once said that ï¿½the only worse design than a pie chart is several of them, for then the viewer is asked to compare quantities located in spatial disarray both within and between piesï¿½ (Edward Tufte, The Visual Display of Quantitative Information, Graphics Press, 1983, p. 178.)

 We do not claim to settle the question - which is multi-facetted and will not have a clear 'winning' design but instead very much depends on the purpose of the chart and task at hand.

\section{Results and Discussion}

\section{Conclusion}

%% if specified like this the section will be ommitted in review mode
\acknowledgments{
This work was supported in part by
NSF DMS 1007697.}

\bibliographystyle{abbrv}
%%use following if all content of bibtex file should be shown
%\nocite{*}
\bibliography{references}
\end{document}

### 2012-03-24 code/code-for-figures (Autosaved).R
code for figures

@@ -1,0 +1,37 @@
+cors <- replicate(1000,
+{
+	x = rexp(100)
+	y = rexp(100)
+	cor(x,y)
+}
+)
+
+qplot(geom="histogram", cors)
+
+x <- rexp(100)
+y <- rexp(100)
+cor(x,y)
+qplot(x,y)
+ggsave("rexp.pdf", height=3, width=3)
+
+x <- factor(rbinom(100,1,prob=0.1))
+y <- rexp(100, rate=3)
+t.test(y~x)
+#y <- rbeta(100, shape1=0.25, shape2=0.25)
+qplot(x,y, geom="boxplot")
+ggsave("two-exp.pdf", height=3, width=3)
+write.csv(data.frame(x=x, y=y), file="two-exp.csv", row.names=F)
+pvals <- replicate(1000,
+{
+x <- factor(rbinom(100,1,prob=0.1))
+y <- rexp(100, rate=3)
+t.test(log(y)~x)$p.value
+}
+)
+qplot(pvals, geom="histogram", binwidth=0.05)
+
+
+x <- c(rnorm(50, mean=3), rnorm(50, mean=-3))
+y <- c(rnorm(50, mean=3), rnorm(50, mean=-3))[sample(100,100, FALSE)]
+qplot(x,y)
+ggsave("rnorm.pdf", height=3, width=3)

### 2012-03-24 code/code-for-figures.R

@@ -1,0 +1,21 @@
+cors <- replicate(1000,
+{
+	x = rexp(100)
+	y = rexp(100)
+	cor(x,y)
+}
+)
+
+qplot(geom="histogram", cors)
+
+x <- rexp(100)
+y <- rexp(100)
+cor(x,y)
+qplot(x,y)
+ggsave("rexp.pdf", height=3, width=3)
+
+
+x <- c(rnorm(50, mean=-3), rnorm(50, mean=3))
+
+y <- c(rnorm(50, mean=3), rnorm(50, mean=-3))[sample(100,100, FALSE)]
+qplot(x,y)

### 2012-03-24 comparison.tex
breaking into separate files

@@
 
-%% \section{Introduction} %for journal use above \firstsection{..} instead
-%We introduce and discuss a tool for comparing different charts that are renderings of the same data in order to evaluate them for efficiency.
-Graphics have a central role in the discovery process as well as in the presentation of results. 
-As designer of  charts one of our main goals is to display data so that the information it contains is relayed in the most efficient way. 
-In the search for the best display out of the seemingly endless multitude of methods we are guided by experience, knowledge of perceptual strengths and weaknesses, as well as aesthetic preferences.
-It is hard, if not impossible, to make the final decision on the design in an  objective manner.
-We can make this process less subjective by conducting user studies to evaluate competing designs of the same data, and measure designs for their efficiency by recordung how accurately and quickly viewers can extract pieces of information.
-User studies for evaluating different designs has a long standing tradition.
-For  statistical graphics user studies have their origin in the (convenience) study on evaluating difficulty of a set of different visual tasks by \citet{cleveland:1984}, which has recently been repeated finding similar results based on a larger and more representative study by \citet{kosara:2010}.
+\input{intro}
 
-Usually, we need simulation studies to be able to completely control for the signal strength in the data. 
-For charts, this is not a very realistic scenario.
-The need for displaying information particularly efficiently or accurately usually arises from  a particular problem while analyzing a data set, from which we want to learn new insight.
-Often, it is difficult to simulate data in such a way that it realistically represents the particular problem, which in turn makes it dubious whether the results from a simulation study are actually applicable to the problem at hand, reminding us painfully of George Box's famous quote on models: ``essentially, all models are wrong, but some are useful." \citep[page=424]{box}
-
-Lineup tests  \citep{buja:2009} allow us to directly use the data at hand to evaluate different competing designs by using permutations tests \cite{good:2011} of the original dataset.
-We can therefore get valid quantitative measures for comparing different designs without having to set up a simulation.
-
-When we are presenting a chart we  want to convince our audience of the presence of a particular feature or relationship in the data. 
-In order to enable us to draw a conclusion about presence or absence of a feature in the data, we have, however, to assess what features show up based purely on marginal distributions rather than joint distributions.
-
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-   \includegraphics[height=0.32\linewidth]{rexp.pdf} 
-   \includegraphics[height=0.32\linewidth]{rnorm.pdf} 
-   \includegraphics[height=0.32\linewidth]{two-exp.pdf} 
-   \caption{Examples of charts with distinctive features but independent variables.
-From left to right: scatterplot of $X$ and $Y$ with independent $X, Y \sim Exp_{\lambda=1}$; scatterplot of two variables with mixed normal distribution $N(3,1)$ and $N(-3,1)$; boxplot of $X$ and $Y$ with $X \sim B_{n=100, p=0.1}$ and $Y \sim Exp_{\lambda=1}$.
-}
-   \label{random}
-\end{figure}
-
-Features in the marginal distributions, such as multi-modality, skewness  or outliers, might interfere with our intuition of what `randomness' might mean in a particular situation and lead us to wrong conclusions.
-All of the plots in figure \ref{random} show distinctive features.
-None, of these, however, is a feature of the joint distribution, i.e.
-all of the variables involved are independent of each other.
-The scatterplot on the left of figure \ref{random} shows data sampled from an exponential distribution with rate parameter $\lambda=1$.
-The data in the middle comes from mixing two normal distributions with different means: half of the points has mean $\mu_1 = -3$, the other half has mean $\mu_2=3$, which results in the distinctive four groups.
-The two boxplots on the right appear to show a median shift between the two groups.
-This, however is a result of both the group size (only 7\% of the points is in the second group) and a skew distribution in vertical direction, with $y \sim Exp_{\lambda=1}$.
-
-Lineup plots have been shown as an effective tool \cite{wickham:2010} for enabling us to quantitively assess the strength of the relationship between $X$ and $Y$.
-In a lineup plot the plot of the actual data is -- as in a police line-up -- inserted among decoy plots.
-The decoy or {\it null} plots are generated in a way that is consistent with the null hypothesis, i.e.
-these are plots in which we can assume that any relationship visible is there purely by chance.
-This gives us a reference framework against which to compare the data plot and, if we are still able to identify the data plot from the lineup because of a distinguishing feature, we have formal evidence that the feature in the data is, in fact, not just coincidental.
-
-The strength of this evidence is determined by both the number of independent evaluations as well as the size of the lineup.
-(XXXX needs more motivation HH) The signal strength \cite{majumder:2011} of a particular lineup is given as:
-\[
-\text{signal strength} = 1- \left( \frac{x}{n}\right)^{1/(m-1)},
-\]
-where $m$ is the size of  the lineup, i.e.
-the number of individuals shown, $x$ is the number of times viewers identified the data plot correctly in  $n$ independent evaluations.
-Signal strength is limited to a range of values between 0 and 1.
-Higher values indicate stronger signal.
- In the case of a parametric situation (i.e.
-the feature shown in the chart can be described in parametric form as a real-valued vector $\theta$ in the data), the signal strength is equivalent to the statistical power of the test.
-
-\section{Lineups for Comparing Designs}
-We are going to make use of the signal strength gained from multiple viewings of a lineup in order to evaluate competing designs as follows:
-\begin{enumerate}
-\item{{\bf Create Lineup Data:} create $m-1$ permutations of $y$}
-\item{{\bf Create lineups from competing designs:} use the same data}
-\item{{\bf Evaluate Lineups:} assess both signal strength and time taken by individuals to come to a decision.}
-\end{enumerate}
-differences in signal strength are due to rendering (after correcting for individuals' visual ability)
-
-XXXX should we include nullabor code?
-
-%We also know that humans learn about new information through a visual search pattern that emphasizes new information (Wolfe, 2002; Rensick, et al., 1997).
-Of course, the type of visualization should take the natural human inclinations into account as well.
-For instance, a good visualization will account for both the ãgestalt?
-like perceptual tendencies as well as the detail-orientated nature of cognition (Kosslyn, Thompson and Ganis, 2002; Pylshyn,2003)
-
-\section{Experimental Setup}
-Describe the experimental setup: polar charts versus euclidean - with and without reference lines.
-Figure showing all four types using the same dataset. 
-Discuss construction of each one. Discuss construction of null chart.
-
-The goal of this experiment is to understand the efficiency of displaying information using euclidian and polar coordinates.
-Efficiency is the deviation from independence, which can be simulated by taking permutations of the original dataset.
-Data concerning wind patterns and time between flights at SEA airport is used for these charts.
-The lineup chart is composed of 20 graphics with all of the exact same characteristics except for the data.
-One of the 20 graphics will be using original airport data while the other 19 will be permutations of that data.
-Efficiency of the chart types can be inferred by how accurately individuals pick out the plot with the original data.
-There are other variables which are introduced to further our understanding of the way we perceive euclidian and polar coordinates.
-All combinations of each level of the variables are involved in the experiment for both polar and euclidian coordinates.
-The graphics in each chart will all have exactly the same level of the variables with the only difference being the data.
-
-The other variables which are monitored are existence of reference line, placement of x-axis, and sample size.
-It is possible that one of the chart types benefits more from additional components such as reference lines.
-The polar and euclidian null charts will be shown either with or without a white reference line.
-The line is placed at slightly over 50 percent of the y axis.
-Additionally, the x-axis has four placements.
-The original dataset contains a characteristic wave pattern, which for euclidian coordinates, may be a large factor contributing to the perceived effectiveness of the graphic.
-Characteristic patterns such as the wave pattern may be perceived in a different way when viewed in polar coordinates.
-When the x-axis is is shifted the look of the pattern changes.
-The wave pattern is broken up for euclidian coordinates but for polar coordinates is only shifted around the axis.
-Finally, the amount of information necessary for detection as well as correctness is a measure for efficiency of the charts so the sample size which was taken from the original dataset is of interest.
-The original data set contained XXX,XXX observations from which 2, 4, 6, 8, 10, and 24 percent of the observations were sampled.
-Small sample sizes are easier to obtain because of time and money constraints.
-It would be nice to know if there is one type of chart which can show data trends using less information.
-
- All combinations of these variables were repeated for plots with a reference line and without a reference line, as shown in figure \ref{layouts}.This figure shows examples of graphics showing the actual data, which in the experiment would be randomly positioned among 19 null plots.
-In total there are XXX unique charts which represent all different combinations of sample size, coordinate type, x-axis placement, and whether or not there is a reference line.
-The charts are all assigned a difficulty level, which is based on sample size, on a scale from zero to four.
-There are two plots assigned a difficulty level of zero.
-These two charts, one polar and one Euclidian, were made from computer generated data and intended to be at the level where an individual, if he or she is not purely guessing, should be able to find the correct plot.
-
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-   \includegraphics[width=0.4\linewidth]{Euclidian_NoLine.pdf} \hspace{0.05\linewidth}
-   \includegraphics[width=0.4\linewidth]{Euclidian_Line.pdf} \\
-   \includegraphics[width=0.475\linewidth]{Polar_NoLine.pdf} 
-   \includegraphics[width=0.475\linewidth]{Polar_Line.pdf}  
-   \caption{All four layouts of charts The first row shows the data in polar coordinates and the second row shows the data in euclidian coordinates.
-The first column in both rows show the plots without any reference lines.
-The second column in both rows have a white reference line added.
-}
-   \label{layouts}
-\end{figure}
-
-Amazon's Mechanical Turk is an online service in which individuals can request and perform simple tasks and surveys for money.
-Using Turk we were able to gather data on how people responded to the charts.
-Each person is shown a series of ten charts and asked to choose the graphic on each chart which they think is "different".
-They then identify, from a list of choices, why they chose this particular chart as well as how confident they are on a scale of one to five.
-Personal information such as age group, gender, and education is also collected on a voluntary basis.
-The amount of time it took the individual to answer is also recorded.
-This study allows us to make informed judgments about how the plot type effects efficiency in terms of accuracy and time spent reading the plots.
-Xcite stephen few?X
-
-Effect is deviation from independence. Cannot be properly measured statistically except in highly aggregated situations, where the effect might be washed out.
-
-Most generally this experiment can be paralleled to the pie chart versus barchart discussion.
-The Euclidian coordinate plot is made up of bars *whose?* areas are colored proportionally according to the data.
-The polar coordinate plot is made up of pie slices which are also colored proportionally according to the data.
-Viewers of these charts will visually decode the areas of the bars and pie slices and make a judgement based on their decoding.
-How accurate this judgment is is left to be discussed.
-According to (Kosslyn, Stephen, Graph Design for the Eye and Mind, Oxford University Press, 2006, p.
-40), area is usually perceived as a function of what the actual area is.
-This function is the area raised to an exponent of approximately 0.8 and then multiplied by a constant.
-However, when bars are parallel, the relative height of these bars is perceived very accurately.
-From this information, one could infer that the Euclidian coordinate plot may gain efficiency based on the height of the colored bars.
-
-Pie versus Barchart discussion is pretty old: cite some of the literature:
-
-From Stephen Few, perceptual edge:
-	When a graph is made, quantitative and categorical information is encoded by a display method.
-Then the information is visually decoded.
-This visual perception is a vital link.
-No matter how clever the choice of the information, and no matter how technologically impressive the encoding, a visualization fails if the decoding fails.
-Some display methods lead to efficient, accurate decoding, and others lead to inefficient, inaccurate decoding.
-It is only through scientific study of visual perception that informed judgments can be made about display methods.
-(William S.
-Cleveland, The Elements of Graphing Data, Hobart Press, 1994, p.
-1)
-
-	The systematic distortion of area is captured by ï¿½Stevenï¿½s Power Law,ï¿½ which states that the psychological impression is a function of the actual physical magnitude raised to an exponent (and multiplied by a scaling constant).
-To be precise, the perceived area is usually equal to the actual area raised to an exponent of about 0.8, times a scaling constant...In contrast, relative line length [such as the lengths of bars] is perceived almost perfectly, provided that the lines are oriented the same way.
-(Kosslyn, Stephen, Graph Design for the Eye and Mind, Oxford University Press, 2006, p.
-40)
-
-	We make angle judgments when we read a pie chart, but we donï¿½t judge angles very well.
-These judgments are biased; we underestimate acute angles (angles less than 90ï¿½) and overestimate obtuse angles (angles greater than 90ï¿½).
-Also, angles with horizontal bisectors (when the line dividing the angle in two is horizontal) appear larger than angles with vertical bisectors.
-(Naomi Robbins, Creating More Effective Graphs, Wiley, 2005, p.
-49)
-
-	 Edward Tufte once said that ï¿½the only worse design than a pie chart is several of them, for then the viewer is asked to compare quantities located in spatial disarray both within and between piesï¿½ (Edward Tufte, The Visual Display of Quantitative Information, Graphics Press, 1983, p.
-178.)
-
- We do not claim to settle the question - which is multi-facetted and will not have a clear 'winning' design but instead very much depends on the purpose of the chart and task at hand.
-
-\section{Results and Discussion}
+\input{experiment}
 

### 2012-03-24 intro.tex
breaking into separate files

%% \section{Introduction} %for journal use above \firstsection{..} instead
%We introduce and discuss a tool for comparing different charts that are renderings of the same data in order to evaluate them for efficiency.
Graphics have a central role in the discovery process as well as in the presentation of results. 
As designer of  charts one of our main goals is to display data so that the information it contains is relayed in the most efficient way. 
In the search for the best display out of the seemingly endless multitude of methods we are guided by experience, knowledge of perceptual strengths and weaknesses, as well as aesthetic preferences. It is hard, if not impossible, to make the final decision on the design in an  objective manner. 
We can make this process less subjective by conducting user studies to evaluate competing designs of the same data, and measure designs for their efficiency by recordung how accurately and quickly viewers can extract pieces of information. 
User studies for evaluating different designs has a long standing tradition. For  statistical graphics user studies have their origin in the (convenience) study on evaluating difficulty of a set of different visual tasks by \citet{cleveland:1984}, which has recently been repeated finding similar results based on a larger and more representative study by \citet{kosara:2010}.

Usually, we need simulation studies to be able to completely control for the signal strength in the data. 
For charts, this is not a very realistic scenario. The need for displaying information particularly efficiently or accurately usually arises from  a particular problem while analyzing a data set, from which we want to learn new insight. Often, it is difficult to simulate data in such a way that it realistically represents the particular problem, which in turn makes it dubious whether the results from a simulation study are actually applicable to the problem at hand, reminding us painfully of George Box's famous quote on models: ``essentially, all models are wrong, but some are useful." \citep[page=424]{box}

Lineup tests  \citep{buja:2009} allow us to directly use the data at hand to evaluate different competing designs by using permutations tests \cite{good:2011} of the original dataset. We can therefore get valid quantitative measures for comparing different designs without having to set up a simulation.

When we are presenting a chart we  want to convince our audience of the presence of a particular feature or relationship in the data. 
In order to enable us to draw a conclusion about presence or absence of a feature in the data, we have, however, to assess what features show up based purely on marginal distributions rather than joint distributions.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[height=0.32\linewidth]{rexp.pdf} 
   \includegraphics[height=0.32\linewidth]{rnorm.pdf} 
   \includegraphics[height=0.32\linewidth]{two-exp.pdf} 
   \caption{Examples of charts with distinctive features but independent variables. From left to right: scatterplot of $X$ and $Y$ with independent $X, Y \sim Exp_{\lambda=1}$; scatterplot of two variables with mixed normal distribution $N(3,1)$ and $N(-3,1)$; boxplot of $X$ and $Y$ with $X \sim B_{n=100, p=0.1}$ and $Y \sim Exp_{\lambda=1}$. }
   \label{random}
\end{figure}

Features in the marginal distributions, such as multi-modality, skewness  or outliers, might interfere with our intuition of what `randomness' might mean in a particular situation and lead us to wrong conclusions. All of the plots in figure \ref{random} show distinctive features. None, of these, however, is a feature of the joint distribution, i.e. all of the variables involved are independent of each other. The scatterplot on the left of figure \ref{random} shows data sampled from an exponential distribution with rate parameter $\lambda=1$. The data in the middle comes from mixing two normal distributions with different means: half of the points has mean $\mu_1 = -3$, the other half has mean $\mu_2=3$, which results in the distinctive four groups.
The two boxplots on the right appear to show a median shift between the two groups. This, however is a result of both the group size (only 7\% of the points is in the second group) and a skew distribution in vertical direction, with $y \sim Exp_{\lambda=1}$.

Lineup plots have been shown as an effective tool \cite{wickham:2010} for enabling us to quantitively assess the strength of the relationship between $X$ and $Y$.
In a lineup plot the plot of the actual data is -- as in a police line-up -- inserted among decoy plots. The decoy or {\it null} plots are generated in a way that is consistent with the null hypothesis, i.e. these are plots in which we can assume that any relationship visible is there purely by chance. This gives us a reference framework against which to compare the data plot and, if we are still able to identify the data plot from the lineup because of a distinguishing feature, we have formal evidence that the feature in the data is, in fact, not just coincidental.

The strength of this evidence is determined by both the number of independent evaluations as well as the size of the lineup. (XXXX needs more motivation HH) The signal strength \cite{majumder:2011} of a particular lineup is given as:
\[
\text{signal strength} = 1- \left( \frac{x}{n}\right)^{1/(m-1)},
\]
where $m$ is the size of  the lineup, i.e. the number of individuals shown, and $x/n$ is the ratio of correctly identified data plots ($x$) in  $n$ independent evaluations. Signal strength is limited to a range of values between 0 and 1. Higher values indicate stronger signal.
 In the case of a parametric situation (i.e. the feature shown in the chart can be described in parametric form as a real-valued vector $\theta$ in the data), the signal strength is equivalent to the statistical power of the test.

\section{Lineups for Comparing Designs}
We are going to make use of the signal strength gained from multiple viewings of a lineup in order to evaluate competing designs as follows:
\begin{enumerate}
\item{{\bf Create Lineup Data:} create $m-1$ permutations of $y$}
\item{{\bf Create lineups from competing designs:} use the same data}
\item{{\bf Evaluate Lineups:} assess both signal strength and time taken by individuals to come to a decision.}
\item{{\bf Evaluate Competing Designs:} differences in signal strength or time to decision are due to differences in the design. In the case of multiple observations for each individual, we can also correct for an individual's visual ability. }
\end{enumerate}

XXXX should we include nullabor code?

%We also know that humans learn about new information through a visual search pattern that emphasizes new information (Wolfe, 2002; Rensick, et al., 1997). Of course, the type of visualization should take the natural human inclinations into account as well. For instance, a good visualization will account for both the ãgestalt? like perceptual tendencies as well as the detail-orientated nature of cognition (Kosslyn, Thompson and Ganis, 2002; Pylshyn,2003)

### 2012-03-24 intro.tex
discussion section

@@
+% !TEX root = comparison.tex
 %% \section{Introduction} %for journal use above \firstsection{..} instead
@@
 It is hard, if not impossible, to make the final decision on the design in an  objective manner.
-We can make this process less subjective by conducting user studies to evaluate competing designs of the same data, and measure designs for their efficiency by recordung how accurately and quickly viewers can extract pieces of information.
+We can make this process less subjective by conducting user studies to evaluate competing designs of the same data, and measure designs for their efficiency by recording how accurately and quickly viewers can extract pieces of information.
 User studies for evaluating different designs has a long standing tradition.
@@
 \begin{enumerate}
-\item{{\bf Create Lineup Data:} create $m-1$ permutations of $y$}
-\item{{\bf Create lineups from competing designs:} use the same data}
-\item{{\bf Evaluate Lineups:} assess both signal strength and time taken by individuals to come to a decision.}
+\item{{\bf Create Lineup Data:} create data for a lineup of size $m$ by  creating  $m-1$ permutations of $y$ or, in the case of a simulation study, drawing $m-1$ samples of size $n$ (the number of rows in the data) from the null distribution.
+Add the original data to the lineup data randomly between 1 and $m$.
+}
+\item{{\bf Create lineups from competing designs:} using the same data, render lineups of all competing designs. }
+\item{{\bf Evaluate Lineups:} by presenting the lineups to independent observers.
+Assess both signal strength and time needed by individuals to come to a decision.
+Note that each observer should only be exposed to each lineup data once.}
 \item{{\bf Evaluate Competing Designs:} differences in signal strength or time to decision are due to differences in the design.
-In the case of multiple observations for each individual, we can also correct for an individual's visual ability.
+In the case that individuals were shown multiple lineups (as part of a bigger study), it is possible to correct outcome measurements for an individual's visual ability.
 }

### 2012-03-24 code/code-for-figures (Autosaved).R

@@ -1,37 +1,0 @@
-cors <- replicate(1000,
-{
-	x = rexp(100)
-	y = rexp(100)
-	cor(x,y)
-}
-)
-
-qplot(geom="histogram", cors)
-
-x <- rexp(100)
-y <- rexp(100)
-cor(x,y)
-qplot(x,y)
-ggsave("rexp.pdf", height=3, width=3)
-
-x <- factor(rbinom(100,1,prob=0.1))
-y <- rexp(100, rate=3)
-t.test(y~x)
-#y <- rbeta(100, shape1=0.25, shape2=0.25)
-qplot(x,y, geom="boxplot")
-ggsave("two-exp.pdf", height=3, width=3)
-write.csv(data.frame(x=x, y=y), file="two-exp.csv", row.names=F)
-pvals <- replicate(1000,
-{
-x <- factor(rbinom(100,1,prob=0.1))
-y <- rexp(100, rate=3)
-t.test(log(y)~x)$p.value
-}
-)
-qplot(pvals, geom="histogram", binwidth=0.05)
-
-
-x <- c(rnorm(50, mean=3), rnorm(50, mean=-3))
-y <- c(rnorm(50, mean=3), rnorm(50, mean=-3))[sample(100,100, FALSE)]
-qplot(x,y)
-ggsave("rnorm.pdf", height=3, width=3)

### 2012-03-24 code/code-for-figures.R

@@ -14,8 +14,24 @@ cor(x,y)
 qplot(x,y)
 ggsave("rexp.pdf", height=3, width=3)
 
+x <- factor(rbinom(100,1,prob=0.1))
+y <- rexp(100, rate=3)
+t.test(y~x)
+#y <- rbeta(100, shape1=0.25, shape2=0.25)
+qplot(x,y, geom="boxplot")
+ggsave("two-exp.pdf", height=3, width=3)
+write.csv(data.frame(x=x, y=y), file="two-exp.csv", row.names=F)
+pvals <- replicate(1000,
+{
+x <- factor(rbinom(100,1,prob=0.1))
+y <- rexp(100, rate=3)
+t.test(log(y)~x)$p.value
+}
+)
+qplot(pvals, geom="histogram", binwidth=0.05)
 
-x <- c(rnorm(50, mean=-3), rnorm(50, mean=3))
 
+x <- c(rnorm(50, mean=3), rnorm(50, mean=-3))
 y <- c(rnorm(50, mean=3), rnorm(50, mean=-3))[sample(100,100, FALSE)]
 qplot(x,y)
+ggsave("rnorm.pdf", height=3, width=3)

### 2012-03-25 intro.tex
results

@@
 User studies for evaluating different designs has a long standing tradition.
-For  statistical graphics user studies have their origin in the (convenience) study on evaluating difficulty of a set of different visual tasks by \citet{cleveland:1984}, which has recently been repeated finding similar results based on a larger and more representative study by \citet{kosara:2010}.
+For  statistical graphics user studies have their origin in the (convenience) study on evaluating difficulty of a set of different visual tasks by \citet{cleveland:1984}, which has recently been replicated with similar results based on a large MTurk study by \citet{heer:2010}.
 
@@
 The strength of this evidence is determined by both the number of independent evaluations as well as the size of the lineup. 
-(XXXX needs more motivation HH) The signal strength \cite{majumder:2011} of a particular lineup is given as:
+Assume a lineup of size $m$ is evaluated by $n$ independent observers, out of which $x$ identify the data plot correctly, which we can use as an estimate for the probability to correctly identify the data plot.
+Correctly identifying the data plot also depends on the features visible in the other plots.
+We can think of a lineup as $m-1$ head-to-head comparisons of the dataplot and $m-1$ nullplots, which the dataplot all has to win, in order to be picked as the plot with the most distinct features.
+This introduces the signal strength \cite{majumder:2011} of a particular lineup as:
 \[
@@
 the feature shown in the chart can be described in parametric form as a real-valued vector $\theta$ in the data), the signal strength is equivalent to the statistical power of the test.
+While we can compute 95\% confidence intervals for signal strength by using 95\% confidence intervals for $x/n$ given as $x/n \pm 1.96 \cdot \sqrt{x/n \cdot (1-x/n)}$ limited to the interval [0,1], the resulting interval is generally to large to be useful.
 

### 2012-03-25 code/code-for-figures.R

@@ -35,3 +35,6 @@ x <- c(rnorm(50, mean=3), rnorm(50, mean=-3))
 y <- c(rnorm(50, mean=3), rnorm(50, mean=-3))[sample(100,100, FALSE)]
 qplot(x,y)
 ggsave("rnorm.pdf", height=3, width=3)
+
+
+x <- rep(1:7, c(1,2,3,4,3,2,1))

### 2012-03-25 intro.tex
intro to power

@@
 Correctly identifying the data plot also depends on the features visible in the other plots.
-We can think of a lineup as $m-1$ head-to-head comparisons of the dataplot and $m-1$ nullplots, which the dataplot all has to win, in order to be picked as the plot with the most distinct features.
-This introduces the signal strength \cite{majumder:2011} of a particular lineup as:
+We can think of a lineup as $m-1$ head-to-head comparisons of the dataplot and $m-1$ nullplots, which the dataplot all has to `win', in order to be picked as the plot with the most distinct features.
+Assuming that observers identify the plot with the stronger signal, the data plot is picked, if it has the strongest signal, leading us to a Beta distribution for the maximum.
+This allows us to introduce the {\it signal strength} \cite{majumder:2011} of a lineup as:
 \[
@@
  In the case of a parametric situation (i.e.
-the feature shown in the chart can be described in parametric form as a real-valued vector $\theta$ in the data), the signal strength is equivalent to the statistical power of the test.
+the feature shown in the chart can be described in parametric form as a real-valued vector $\theta$ in the data), it can be shown \cite{majumder:2010} that the signal strength is approximately linear to the statistical power of the test.
 While we can compute 95\% confidence intervals for signal strength by using 95\% confidence intervals for $x/n$ given as $x/n \pm 1.96 \cdot \sqrt{x/n \cdot (1-x/n)}$ limited to the interval [0,1], the resulting interval is generally to large to be useful.
@@
 
+The power of a statistical test is defined as the probability to reject the null hypothesis.
+Translated to the visual tests, power is the probability to identify the data plot from the lineup.
+While we can compute power only in the parametric setting of a simulation study, we can make use of the signal strength of a lineup.
+
+Power consideration: what is statistical power, how can we assess it.
+
 %We also know that humans learn about new information through a visual search pattern that emphasizes new information (Wolfe, 2002; Rensick, et al., 1997).

### 2012-03-25 intro.tex
changed cls file to refer to pdf not eps, and spelling of euclidian in experiment.tex

@@
 The need for displaying information particularly efficiently or accurately usually arises from  a particular problem while analyzing a data set, from which we want to learn new insight.
-Often, it is difficult to simulate data in such a way that it realistically represents the particular problem, which in turn makes it dubious whether the results from a simulation study are actually applicable to the problem at hand, reminding us painfully of George Box's famous quote on models: ``essentially, all models are wrong, but some are useful." \citep[page=424]{box}
+Often, it is difficult to simulate data in such a way that it realistically represents the particular problem, which in turn makes it dubious whether the results from a simulation study are actually applicable to the problem at hand, reminding us painfully of George Box's famous quote on models: ``essentially, all models are wrong, but some are useful.'' \citep[page=424]{box}
 

### 2012-03-26 intro.tex
intro

@@
 The strength of this evidence is determined by both the number of independent evaluations as well as the size of the lineup. 
-Assume a lineup of size $m$ is evaluated by $n$ independent observers, out of which $x$ identify the data plot correctly, which we can use as an estimate for the probability to correctly identify the data plot.
-Correctly identifying the data plot also depends on the features visible in the other plots.
+Assume a lineup of size $m$ is evaluated by $n$ independent observers, out of which $x$ identify the data plot correctly, which we can use as an estimate for the probability to correctly identify the data plot:
+this leads us to an estimate for the probability of correctly identifying the plot of $x/n$, with 95\% confidence intervals  for $x/n$ given as $x/n \pm 1.96 \cdot \sqrt{x/n \cdot (1-x/n)}$ limited to the interval [0,1].
+Note, that the resulting interval is often too large to be useful.
+
 We can think of a lineup as $m-1$ head-to-head comparisons of the dataplot and $m-1$ nullplots, which the dataplot all has to `win', in order to be picked as the plot with the most distinct features.
-Assuming that observers identify the plot with the stronger signal, the data plot is picked, if it has the strongest signal, leading us to a Beta distribution for the maximum.
-This allows us to introduce the {\it signal strength} \cite{majumder:2011} of a lineup as:
+Assuming that observers identify the plot with the strongest signal, the probability that data plot is picked follows a Beta distribution for the maximum:
+this allows us to introduce the {\it signal strength} \cite{majumder:2011} of a lineup as:
 \[
-\text{signal strength} = 1- \left( \frac{x}{n}\right)^{1/(m-1)},
+\text{signal strength} =  \left( \frac{x}{n}\right)^{1/(m-1)}.
 \]
-where $m$ is the size of  the lineup, i.e.
-the number of individuals shown, and $x/n$ is the ratio of correctly identified data plots ($x$) in  $n$ independent evaluations.
-Signal strength is limited to a range of values between 0 and 1.
-Higher values indicate stronger signal.
+Signal strength is limited to a range of values between 0 and 1. Higher values indicate stronger signal.
  In the case of a parametric situation (i.e.
-the feature shown in the chart can be described in parametric form as a real-valued vector $\theta$ in the data), it can be shown \cite{majumder:2010} that the signal strength is approximately linear to the statistical power of the test.
-While we can compute 95\% confidence intervals for signal strength by using 95\% confidence intervals for $x/n$ given as $x/n \pm 1.96 \cdot \sqrt{x/n \cdot (1-x/n)}$ limited to the interval [0,1], the resulting interval is generally to large to be useful.
+the feature shown in the chart can be described in parametric form as a real-valued vector $\theta$ in the data), it can be shown \cite{majumder:2011} that the signal strength is strongly correlated to the $p$-value of the corresponding statistical test for $\theta$.
 
@@
 
-The power of a statistical test is defined as the probability to reject the null hypothesis.
-Translated to the visual tests, power is the probability to identify the data plot from the lineup.
-While we can compute power only in the parametric setting of a simulation study, we can make use of the signal strength of a lineup.
+The power of a statistical test is defined as the probability to reject the null hypothesis. Translated to the visual tests, power is the probability to identify the data plot from the lineup. 
+Comparing power of competing designs therefore involves comparing percentages of correct responses $p_1$ and $p_2$. An $\alpha\cdot$100\% confidence interval for this comparison is given as 
+\[
+p_1 - p_2 \pm t_{1-\alpha/2, n-1} \sqrt{p_1(1-p_1)/n_1 + p2(1-p_2)/n_2},
+\]
+where $n$ is Welch-Satterthwaite \cite{welch:1947} of the degrees of freedom. Note that we use $p_i = (x_i+1)/(n_i+1)$ and $n_i+2$ for a better coverage of the confidence interval \cite{agresti:1998}.
 
-Power consideration: what is statistical power, how can we assess it.
+\subsection{First Example}
 

### 2012-03-26 intro.tex
lineup comparisons

@@
 \[
-p_1 - p_2 \pm t_{1-\alpha/2, n-1} \sqrt{p_1(1-p_1)/n_1 + p2(1-p_2)/n_2},
+p_1 - p_2 \pm t_{1-\alpha/2, n-1} \sqrt{p_1(1-p_1)/n_1 + p_2(1-p_2)/n_2},
 \]
@@
 
+While this allows a direct comparison of the designs, we cannot adjust for the individuals' perceptual abilities.
+In the case that we have multiple responses from each person (i.e.
+several {\it different} lineup tasks), we can estimate for their perceptual ability and correct power differences between competing designs accordingly, e.g.
+by modeling their power with a subject-specific random intercept.
+
 \subsection{First Example}
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+   \includegraphics[width=\linewidth]{turk4-designs.pdf} 
+   \caption{Power results for two competing designs: polar versus a Euclidean, with and without a reference line. Plots are facetted by percentage of data used. }
+   \label{fig:example}
+\end{figure}
 

### 2012-03-26 experiment.tex
re-structuring, focus on first example

\section{Results and Discussion}

Euclidian coordinates performed better when speaking of efficiency in terms of accuracy and speed. 76.92\% of the Euclidian charts shown resulted in the a correct identification of the real data while only 20.31\% of the polar charts did the same. A t-test comparing means for charts with polar coordinates versus charts with Euclidian coordinates resulted in a t statistic of 15.2013 with a p-value of 2.2e-16. Similarly, a t-test comparing means of time spent on charts with Euclidian coordinates versus charts with polar coordinates resulted in a p-value of 1.014e-11. The average of the log of time spent on each individual chart for charts with Euclidian coordinates is 3.53 minutes while it was 4.07 minutes for charts with polar coordinates. Euclidian coordinates resulted in a significantly more accurate identification of the real data set in significantly shorter time.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=3in]{turk4_time_perc_preds.pdf}  
   \caption{Values predicted from XXXX how to say itXXXX of accuracy for log of time taken and test parameter.}
   \label{accuracy_preds}
\end{figure}

Figure \ref{accuracy_preds} shows predicted accuracy levels. These predicted values were obtained by using R package lme4 and a mixed effect model explaining correct responses by type of chart and the log of time spent while accounting for different levels of accuracy for each individual who participated in the experiment. The predicted values show that there is an overall higher accuracy level for Euclidian coordinates when compared to polar coordinates. For both chart types, as the time spent increases, the predicted accuracy decreases.

### 2012-03-26 intro.tex
re-structuring, focus on first example

--- 2012-03-26 intro.tex
@@
 
-\section{Lineups for Comparing Designs}
+\section{Comparison of Designs}
 We are going to make use of the signal strength gained from multiple viewings of a lineup in order to evaluate competing designs as follows:
@@
 \[
-p_1 - p_2 \pm t_{1-\alpha/2, n-1} \sqrt{p_1(1-p_1)/n_1 + p2(1-p_2)/n_2},
+p_1 - p_2 \pm t_{1-\alpha/2, n-1} \sqrt{p_1(1-p_1)/n_1 + p_2(1-p_2)/n_2},
 \]
-where $n$ is Welch-Satterthwaite \cite{welch:1947} of the degrees of freedom. Note that we use $p_i = (x_i+1)/(n_i+1)$ and $n_i+2$ for a better coverage of the confidence interval \cite{agresti:1998}.
+where $n$ is the Welch-Satterthwaite \cite{welch:1947} estimate of the degrees of freedom.
+Note that we use $p_i = (x_i+1)/(n_i+1)$ and $n_i+2$ for a better coverage of the confidence interval \cite{agresti:1998}.
 
-\subsection{First Example}
+While this allows a direct comparison of the designs, we cannot adjust for the individuals' perceptual abilities.
+In the case that we have multiple responses from each person (i.e.
+data is collected on several {\it different} lineup tasks), we can estimate for their perceptual ability and correct power differences between competing designs accordingly, e.g.
+by modeling power using  a subject-specific random intercept in a generalized linear model.
 

### 2012-03-27 intro.tex
Re-worked the first paragraph of the intro

@@
 %We introduce and discuss a tool for comparing different charts that are renderings of the same data in order to evaluate them for efficiency.
-Graphics have a central role in the discovery process as well as in the presentation of results. 
+So, you've just spent countless hours, burning the late night candle
+wax, and discovered several astonishing features in the data. Your
+exploratory graphics are quick to produce, interactive, flexible,
+fabulous for discovery, but now you need to present the information to
+your boss, the co-workers, a class of students, or get it published in
+that top-rated journal. Those exploratory graphics are unlikely to
+communicate the information to your audience efficiently.
+%Graphics have a central role in the discovery process as well as in the presentation of results.
 As designer of  charts one of our main goals is to display data so that the information it contains is relayed in the most efficient way.
-In the search for the best display out of the seemingly endless multitude of methods we are guided by experience, knowledge of perceptual strengths and weaknesses, as well as aesthetic preferences.
-It is hard, if not impossible, to make the final decision on the design in an  objective manner.
+In deciding on the best presentation display %out of the seemingly endless multitude of methods
+we will guided by past experience, aesthetic preferences, and knowledge of general perceptual strengths and weaknesses determined from data graphics user studies such as \citet{cleveland:1984} and the large MTurk study in \citet{heer:2010}.
+But it is hard to take these general principles and apply them to specific visual communication task, especially in an objective manner.
+
+%knowledge of perceptual strengths and weaknesses,  .
+But it is hard to make the final decision on the design in an objective manner.
 We can make this process less subjective by conducting user studies to evaluate competing designs of the same data, and measure designs for their efficiency by recording how accurately and quickly viewers can extract pieces of information.
@@
 For  statistical graphics user studies have their origin in the (convenience) study on evaluating difficulty of a set of different visual tasks by \citet{cleveland:1984}, which has recently been replicated with similar results based on a large MTurk study by \citet{heer:2010}.
+
+%% Di needs to work on the next few paragraphs, too.
 

### 2012-03-27 experiment.tex
experimental setup

@@
 For both chart types, as the time spent increases, the predicted accuracy decreases.
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+   \includegraphics[width=3in]{turk4_samplesize_correctn.pdf}  
+   \caption{For each sample size for polar and Euclidian coordinates, the bars are colored according to the number of correct responses.
+Darker colors indicate lower accuracy levels while lighter colors indicate higher accuracy levels.
+}
+   \label{samp_size_acc}
+\end{figure}
+
+The difficulty of the charts is largely determined by the sample size taken from the original data set.
+Samples of 24 \% generally result in higher accuracy as the extra information facilitates detection of a relationship.
+Figure \ref{accuracy_preds} shows the relationship between the number of correct responses relative to the total number of responses and how this relationship differs between Euclidian and polar coordinates.
+The bars in the Euclidian plot are overall a lighter color than they are in the polar plot.
+This indicates a higher level of accuracy at all sample sizes.
+For Euclidian coordinates, we can see a stair step pattern in the accuracy levels as sample size increases.
+Lower levels of accuracy, indicated by darker colors, become consistently less prevelant as sample size increases.
+The relationship between sample size and accuracy level is not as simple for polar coordinates.
+Sample sizes of 2\% and 4\% have very similar accuracy levels.
+After 6\% the stair step pattern becomes visible and similar to that of Euclidian coordinates.
+Sample size seems to have less of an effect on accuracy for polar coordinates than it does for Euclidian coordinates.

### 2012-03-27 experiment.tex
updated figures

@@
 
+Ideally, efficient detection of data characteristics will be possible at small sample sizes.
 The difficulty of the charts is largely determined by the sample size taken from the original data set.
@@
 Sample size seems to have less of an effect on accuracy for polar coordinates than it does for Euclidian coordinates.
+Predicted values of the proportion of correct responses when explained by sample size, the type of chart, and the interaction between these two factors, show that there is a significant difference between the relationship between sample size and correct responses for polar and euclidian coordinates.
+Indeed, the proportion of correct responses increases more rapidly as sample size increases for Euclidian coordinates than it does for polar coordinates.
+
+The effect of a reference line on accuracy and time was not significant for both types of charts.

### 2012-03-27 code/code-for-figures.R

@@ -1,3 +1,44 @@
+# lineup for airline
+xx <- seq(0, 360, by=1)
+yy <- rep(0.5, length=length(xx))
+refdf <- data.frame(xx=xx, yy=yy)
+
+ggplot(flights_sample, aes(x = (WindDirDegrees-1 + X)%%360, fill = factor(nf, levels = c(0:8)))) +
+geom_bar(binwidth = 10, position="fill") +
+scale_x_continuous(limits = c(0, 360), breaks=seq(0, 360, by=45), labels=c("N", "NE","E","SE","S","SW","W","NW","N")) +
+coord_polar(start = 0) + geom_line(aes(x=xx, y=yy), size = .5, alpha=0.5, colour = "white", inherit.aes=FALSE, data=refdf) + opts(legend.position="none", plot.margin=unit(c(0,0,0,0), "cm")) +
+seq7rev + labs(x = " ", y = " ") 
+ggsave(file="polar_line.pdf", width=4, height=4, dpi=75)
+
+
+ggplot(flights_sample, aes(x = (WindDirDegrees-1 + X)%%360, fill = factor(nf, levels = c(0:8)))) +
+geom_bar(binwidth = 10, position="fill") +
+scale_x_continuous(limits = c(0, 360), breaks=seq(0, 360, by=45), labels=c("N", "NE","E","SE","S","SW","W","NW","N")) +
+coord_polar(start = 0) + opts(legend.position="none", plot.margin=unit(c(0,0,0,0), "cm")) +
+seq7rev + labs(x = " ", y = " ") 
+ggsave(file="polar_noline.pdf", width=4, height=4, dpi=75)
+
+
+ggplot(flights_sample, aes(x = (WindDirDegrees-1 + X)%%360, fill = factor(nf, levels = c(0:8)))) +
+geom_bar(binwidth = 10, position="fill", alpha=0.8) +
+scale_x_continuous(limits = c(0, 360), breaks=seq(0, 360, by=45), labels=c("N", "NE","E","SE","S","SW","W","NW","N")) + geom_hline(yintercept=0.5, size = .5, alpha=0.5, colour = "white") + opts(legend.position="none", plot.margin=unit(c(0,0,0,0), "cm")) +
+seq7rev + labs(x = " ", y = " ") 
+
+ggsave(file="euclid_line.pdf", width=4, height=4, dpi=75)
+
+
[…]

### 2012-03-28 intro.tex
working on intro, added/fixed references

@@
 that top-rated journal. Those exploratory graphics are unlikely to
-communicate the information to your audience efficiently.
-%Graphics have a central role in the discovery process as well as in the presentation of results.
-As designer of  charts one of our main goals is to display data so that the information it contains is relayed in the most efficient way.
-In deciding on the best presentation display %out of the seemingly endless multitude of methods
-we will guided by past experience, aesthetic preferences, and knowledge of general perceptual strengths and weaknesses determined from data graphics user studies such as \citet{cleveland:1984} and the large MTurk study in \citet{heer:2010}.
+communicate the discovered information to your audience efficiently,
+or elegantly.  To decide on the best way to present the specific
+information there can be many types of plots to choose from. The
+decision process is assisted by past experience, personal aesthetic
+preferences, the characteristics of the audience, and a good working
+knowledge of general perceptual strengths and weaknesses
+\citep{cleveland:1984,heer:2010}. Perceptual studies are thinly spread
+for data visualization, and have, and will probably, only cover very
+broad perceptual principles. Faced with design decisions for a very
+specific task requires a closer level of detail. Focus groups
+\citep{Mazza07} can find small problems with designs and move us
+closer to a final handful of possible designs, but they lack broad
+audience validatity and objective evaluation.
+
+%determined from data graphics user studies such as \citet{cleveland:1984} and the large MTurk study in \citet{}.
 But it is hard to take these general principles and apply them to specific visual communication task, especially in an objective manner.

### 2012-03-28 experiment-two.tex
small restructuring

\subsection{Comparing Distributions}

\begin{table}[ht]
\begin{center}
\begin{tabular}{rrrrr}
  \hline
 & Estimate & Std. Error & z value & Pr($>$$|$z$|$) \\ 
  \hline
(Intercept) & -2.00 & 0.38 & -5.29 & 0.00 \\ 
  test\_paramdensity & -0.17 & 0.53 & -0.32 & 0.75 \\ 
  test\_paramdotplot & -0.72 & 0.52 & -1.39 & 0.17 \\ 
  test\_paramhistogram & -0.30 & 0.53 & -0.58 & 0.56 \\ 
  d & 1.77 & 0.34 & 5.19 & 0.00 \\ 
  n1 & 0.02 & 0.00 & 9.80 & 0.00 \\ 
  I(n1/n2) & -0.31 & 0.11 & -2.93 & 0.00 \\ 
  test\_paramdensity:d & -0.67 & 0.47 & -1.42 & 0.15 \\ 
  test\_paramdotplot:d & 0.05 & 0.46 & 0.11 & 0.91 \\ 
  test\_paramhistogram:d & -0.27 & 0.48 & -0.56 & 0.57 \\ 
  test\_paramdensity:n1 & -0.00 & 0.00 & -0.40 & 0.69 \\ 
  test\_paramdotplot:n1 & -0.01 & 0.00 & -3.06 & 0.00 \\ 
  test\_paramhistogram:n1 & -0.01 & 0.00 & -2.67 & 0.01 \\ 
  test\_paramdensity:I(n1/n2) & 0.06 & 0.15 & 0.42 & 0.68 \\ 
  test\_paramdotplot:I(n1/n2) & 0.27 & 0.15 & 1.81 & 0.07 \\ 
  test\_paramhistogram:I(n1/n2) & 0.10 & 0.15 & 0.65 & 0.52 \\ 
   \hline
\end{tabular}
\end{center}
\end{table}

### 2012-03-28 experiment-two.tex
p values in tables

@@
 (Intercept) & -2.00 & 0.38 & -5.29 & 0.00 \\ 
-  test\_paramdensity & -0.17 & 0.53 & -0.32 & 0.75 \\ 
+  test\_paramdensity & -0.29 & 0.53 & -0.55 & 0.58 \\ 
   test\_paramdotplot & -0.72 & 0.52 & -1.39 & 0.17 \\ 
-  test\_paramhistogram & -0.30 & 0.53 & -0.58 & 0.56 \\ 
-  d & 1.77 & 0.34 & 5.19 & 0.00 \\ 
-  n1 & 0.02 & 0.00 & 9.80 & 0.00 \\ 
-  I(n1/n2) & -0.31 & 0.11 & -2.93 & 0.00 \\ 
-  test\_paramdensity:d & -0.67 & 0.47 & -1.42 & 0.15 \\ 
-  test\_paramdotplot:d & 0.05 & 0.46 & 0.11 & 0.91 \\ 
-  test\_paramhistogram:d & -0.27 & 0.48 & -0.56 & 0.57 \\ 
+  test\_paramhistogram & -0.29 & 0.53 & -0.54 & 0.59 \\ 
+  d & 1.78 & 0.34 & 5.22 & 0.00 \\ 
+  n1 & 0.02 & 0.00 & 9.77 & 0.00 \\ 
+  I(n1/n2) & -0.31 & 0.11 & -2.94 & 0.00 \\ 
+  test\_paramdensity:d & -0.61 & 0.47 & -1.29 & 0.20 \\ 
+  test\_paramdotplot:d & 0.04 & 0.46 & 0.08 & 0.94 \\ 
+  test\_paramhistogram:d & -0.29 & 0.48 & -0.60 & 0.55 \\ 
   test\_paramdensity:n1 & -0.00 & 0.00 & -0.40 & 0.69 \\ 
-  test\_paramdotplot:n1 & -0.01 & 0.00 & -3.06 & 0.00 \\ 
+  test\_paramdotplot:n1 & -0.01 & 0.00 & -3.04 & 0.00 \\ 
   test\_paramhistogram:n1 & -0.01 & 0.00 & -2.67 & 0.01 \\ 
-  test\_paramdensity:I(n1/n2) & 0.06 & 0.15 & 0.42 & 0.68 \\ 
+  test\_paramdensity:I(n1/n2) & 0.09 & 0.15 & 0.59 & 0.55 \\ 
   test\_paramdotplot:I(n1/n2) & 0.27 & 0.15 & 1.81 & 0.07 \\ 
-  test\_paramhistogram:I(n1/n2) & 0.10 & 0.15 & 0.65 & 0.52 \\ 
+  test\_paramhistogram:I(n1/n2) & 0.09 & 0.15 & 0.62 & 0.53 \\ 
    \hline
@@
 \end{center}
+\caption{correctness}
 \end{table}
+
+\begin{table}[ht]
+\begin{center}
+\begin{tabular}{rrrrr}
+  \hline
+ & Estimate & Std..Error & t.value & pval \\ 
+  \hline
+(Intercept) & 3.85 & 0.09 & 42.21 & 0.00 \\ 
+  test\_paramdensity & 0.01 & 0.12 & 0.07 & 0.94 \\ 
+  test\_paramdotplot & 0.03 & 0.12 & 0.26 & 0.80 \\ 
+  test\_paramhistogram & 0.23 & 0.12 & 1.97 & 0.05 \\ 
+  d & -0.23 & 0.08 & -3.02 & 0.00 \\ 
+  n1 & 0.00 & 0.00 & 1.74 & 0.08 \\ 
+  I(n1/n2) & -0.04 & 0.02 & -1.48 & 0.14 \\ 
+  test\_paramdensity:d & 0.11 & 0.11 & 1.02 & 0.31 \\ 
+  test\_paramdotplot:d & 0.20 & 0.11 & 1.86 & 0.06 \\ 
+  test\_paramhistogram:d & -0.01 & 0.11 & -0.09 & 0.93 \\ 
+  test\_paramdensity:n1 & -0.00 & 0.00 & -2.15 & 0.03 \\ 
+  test\_paramdotplot:n1 & -0.00 & 0.00 & -2.87 & 0.00 \\ 
+  test\_paramhistogram:n1 & -0.00 & 0.00 & -0.93 & 0.35 \\ 
+  test\_paramdensity:I(n1/n2) & 0.04 & 0.03 & 1.18 & 0.24 \\ 
+  test\_paramdotplot:I(n1/n2) & 0.02 & 0.03 & 0.65 & 0.52 \\ 
+  test\_paramhistogram:I(n1/n2) & 0.00 & 0.04 & 0.04 & 0.97 \\ 
+   \hline
+\end{tabular}
+\end{center}
+\caption{log time taken}
+
+\end{table}

### 2012-03-28 comparison.tex
re-structuring, focus on first example
discussion of first experiment
small restructuring
re-working first paragraph - yep still! added more references.

@@
 \usepackage{amsmath}
-
+\usepackage{multirow}
+\usepackage{url}
+\usepackage[table]{xcolor}
 %% We encourage the use of mathptmx for consistent usage of times font
@@
 %% indicate IEEE Member or Student Member in form indicated below
-\author{Lendie Follett, Heike Hofmann, Mahbub Majumder, Dianne Cook}
+\author{Lendie Follett, Heike Hofmann, Mahbubul Majumder, Dianne Cook}
 \authorfooter{
@@
 \item
- Lendie Follett is a graduate student in Statistics at Iowa State University, E-mail: lfollett@iastate.edu.
+ Lendie Follett is an undergraduate student in Statistics at Iowa State University, E-mail: lfollett@iastate.edu.
 \item
-everybody else is at Iowa State University, too
+ Heike Hofmann is an associate professor in Statistics at Iowa State University, E-mail: hofmann@iastate.edu.
+\item
+ Mahbubul Majumder is a graduate student in Statistics at Iowa State University, E-mail: mahbub@iastate.edu.
+\item
+ Dianne Cook is a full professor in Statistics at Iowa State University, E-mail: dicook@iastate.edu.
 }
@@
 
+\input{setup}
 \input{experiment}
-
-\section{Conclusion}
+\input{experiment-two}
+\input{conclusions}
 

### 2012-03-28 intro.tex
re-working first paragraph - yep still! added more references.

@@
 preferences, the characteristics of the audience, and a good working
-knowledge of general perceptual strengths and weaknesses
-\citep{cleveland:1984,heer:2010}. Perceptual studies are thinly spread
-for data visualization, and have, and will probably, only cover very
-broad perceptual principles. Faced with design decisions for a very
-specific task requires a closer level of detail. Focus groups
-\citep{Mazza07} can find small problems with designs and move us
-closer to a final handful of possible designs, but they lack broad
-audience validatity and objective evaluation.
+knowledge of general perceptual strengths and weaknesses of particular
+displays \citep{cleveland:1984,heer:2010}. Perceptual studies are
+thinly spread for data visualization, and have, and will probably,
+only cover very broad perceptual principles. Faced with design
+decisions for a very specific task requires a closer level of
+detail. Focus groups \citep{Mazza07} can find small problems with
+designs and move us closer to a final handful of possible
+designs. Usability studies and case studies
+(e.g. \citet{Plaisant:2004}, \citet{Shneiderman:2006}) typically focus
+on the use of particular software for solving data exploration tasks.
+These approaches lack broad audience validity and
+objectiveness. Several papers indicate continuing issues arising with
+evaluating information visualizations
+\citep{North:2006,Ellis:2006,Viegas:2006,Munzer:2009}.
 

### 2012-03-28 intro.tex
tweaked start of paper, to integrate literature

@@
 thinly spread for data visualization, and have, and will probably,
-only cover very broad perceptual principles. Faced with design
+only cover very broad perceptual principles, primarily because they
+tend to be based solely on simulated data. In order to study
+perceptual principles in a controlled experimental setting simulated
+data is unavoidable. This provides principles that may not apply
+closely enough to a specific analysis task.  Faced with design
 decisions for a very specific task requires a closer level of
@@
 objectiveness. Several papers indicate continuing issues arising with
-evaluating information visualizations
-\citep{North:2006,Ellis:2006,Viegas:2006,Munzer:2009}.
+and suggestions for improving evaluation of information visualizations
+\citep{North:2006,Ellis:2006,Viegas:2006,Munzer:2009}. Here we suggest
+a new approach using recent research on visual inference.
 
@@
 
+Lineups \citep{buja:2009,wickham:2010} involving real data coupled
+with simulated data will enable both specific plot design evaluation
+and provide support for broad validity. Lineups allow us to directly
+use the data at hand to evaluate different competing designs. Coupling
+this with simulated data studying the feature discovered in the real
+data enables the addition of the calculation of a signal strength
+metric to quantitatively compare the different designs.
+
 Usually, we need simulation studies to be able to completely control for the signal strength in the data. 
@@
 
-Lineup tests  \citep{buja:2009} allow us to directly use the data at hand to evaluate different competing designs by using permutations tests \cite{good:2011} of the original dataset.
+Lineup \citep{buja:2009} allow us to directly use the data at hand to evaluate different competing designs by using permutations tests \cite{good:2011} of the original dataset.
 We can therefore get valid quantitative measures for comparing different designs without having to set up a simulation.

### 2012-03-28 code/code-for-figures.R
images for second experiment

@@ -79,3 +79,53 @@ ggsave("rnorm.pdf", height=3, width=3)
 
 
 x <- rep(1:7, c(1,2,3,4,3,2,1))
+
+
+
+
+
+####################
+# turk 5 pictures
+# conc data:
+conc <- read.csv("/Users/heike/Dropbox/Lineups-nf/turk5/data/turk5_1_15_5_12_16.csv")
+
+sample <- subset(conc, .sample==16)
+t.test(values~group, data=sample)
+
+lineup.ops <- opts(legend.position="none", axis.ticks = theme_blank(), #axis.text.y = theme_blank(), 
+axis.text.x = theme_blank(), plot.margin=unit(c(0,0,0,0), "cm"))
+
+ggplot(aes(x=values,   fill=factor(group)), data=sample) + geom_histogram(binwidth=25, alpha=0.5)+ xlab("")  + ylab("") + lineup.ops
+ggsave("hist-conc.pdf")
+
+alpha=0.5
+ggplot(aes(x=factor(group), y=values, fill=factor(group), colour=factor(group)), data=sample)   + geom_boxplot(alpha=alpha) + coord_flip() + lineup.ops + xlab("")  + ylab("") 
+ggsave("boxplot-conc.pdf")
+
+
+ggplot(aes(x=values,  colour=factor(group), fill=factor(group)), data=sample) + geom_density(alpha=alpha)+ lineup.ops + xlab("")  + ylab("") 
+ggsave("density-conc.pdf")
+
+ggplot(aes(x=values, y=factor(group), colour=factor(group)), data=sample)  + geom_jitter(position=position_jitter(height=0.1, width=0), size=4, alpha=alpha)+ lineup.ops + xlab("")  + ylab("") 
+ggsave("dotplot-conc.pdf")
+
+conc.res <- subturk[grep("_15_15_1_", as.character(subturk$pic_name)),]
+
+stats <- ddply(conc.res, .(n1,n2,d, rep), summarize,
+	correct=sum(response),
+	n=length(param_value),
+	difficulty=mean(difficulty),
+	pval=mean(p_value)
+)
+qplot(interaction(n1,n2,d),correct/n, data=stats) + coord_flip()
+
+conc.res <- subset(subturk, n1==15 & n2==15 & d==1)
+ddply(conc.res, .(n1,n2,d,test_param), summarize,
+	correct=sum(response),
+	n=length(param_value),
[…]

### 2012-03-29 experiment-two.tex
images for second experiment
second experiment

--- 2012-03-28 experiment-two.tex
@@
+% !TEX root = comparison.tex
 \subsection{Comparing Distributions}
+
+\begin{figure*} [hbtp]
+   \centering
+   \includegraphics[width=0.245\textwidth]{hist-conc} 
+   \includegraphics[width=0.245 \textwidth]{boxplot-conc} 
+   \includegraphics[width=0.245 \textwidth]{density-conc} 
+   \includegraphics[width=0.245 \textwidth]{dotplot-conc} 
+   \caption{Overview of all four competing designs for displaying differences in distributions.}
+   \label{fig:expii}
+\end{figure*}
 
@@
 (Intercept) & -2.00 & 0.38 & -5.29 & 0.00 \\ 
-  test\_paramdensity & -0.17 & 0.53 & -0.32 & 0.75 \\ 
-  test\_paramdotplot & -0.72 & 0.52 & -1.39 & 0.17 \\ 
-  test\_paramhistogram & -0.30 & 0.53 & -0.58 & 0.56 \\ 
-  d & 1.77 & 0.34 & 5.19 & 0.00 \\ 
-  n1 & 0.02 & 0.00 & 9.80 & 0.00 \\ 
-  I(n1/n2) & -0.31 & 0.11 & -2.93 & 0.00 \\ 
-  test\_paramdensity:d & -0.67 & 0.47 & -1.42 & 0.15 \\ 
-  test\_paramdotplot:d & 0.05 & 0.46 & 0.11 & 0.91 \\ 
-  test\_paramhistogram:d & -0.27 & 0.48 & -0.56 & 0.57 \\ 
-  test\_paramdensity:n1 & -0.00 & 0.00 & -0.40 & 0.69 \\ 
-  test\_paramdotplot:n1 & -0.01 & 0.00 & -3.06 & 0.00 \\ 
-  test\_paramhistogram:n1 & -0.01 & 0.00 & -2.67 & 0.01 \\ 
-  test\_paramdensity:I(n1/n2) & 0.06 & 0.15 & 0.42 & 0.68 \\ 
-  test\_paramdotplot:I(n1/n2) & 0.27 & 0.15 & 1.81 & 0.07 \\ 
-  test\_paramhistogram:I(n1/n2) & 0.10 & 0.15 & 0.65 & 0.52 \\ 
+  density & -0.29 & 0.53 & -0.55 & 0.58 \\ 
+  dotplot & -0.72 & 0.52 & -1.39 & 0.17 \\ 
+  histogram & -0.29 & 0.53 & -0.54 & 0.59 \\ 
+  d & 1.78 & 0.34 & 5.22 & 0.00 \\ 
+  n1 & 0.02 & 0.00 & 9.77 & 0.00 \\ 
+  I(n1/n2) & -0.31 & 0.11 & -2.94 & 0.00 \\ 
+  density:d & -0.61 & 0.47 & -1.29 & 0.20 \\ 
+  dotplot:d & 0.04 & 0.46 & 0.08 & 0.94 \\ 
+  histogram:d & -0.29 & 0.48 & -0.60 & 0.55 \\ 
+  density:n1 & -0.00 & 0.00 & -0.40 & 0.69 \\ 
+  dotplot:n1 & -0.01 & 0.00 & -3.04 & 0.00 \\ 
+  histogram:n1 & -0.01 & 0.00 & -2.67 & 0.01 \\ 
+  density:I(n1/n2) & 0.09 & 0.15 & 0.59 & 0.55 \\ 
+  dotplot:I(n1/n2) & 0.27 & 0.15 & 1.81 & 0.07 \\ 
+  histogram:I(n1/n2) & 0.09 & 0.15 & 0.62 & 0.53 \\ 
    \hline
@@
 \end{center}
+\caption{correctness}
 \end{table}
+
+\begin{table}[ht]
+\begin{center}
+\begin{tabular}{rrrrr}
+  \hline
+ & Estimate & Std..Error & t.value & pval \\ 
+  \hline
+(Intercept) & 3.85 & 0.09 & 42.21 & 0.00 \\ 
+  density & 0.01 & 0.12 & 0.07 & 0.94 \\ 
+  dotplot & 0.03 & 0.12 & 0.26 & 0.80 \\ 
+  histogram & 0.23 & 0.12 & 1.97 & 0.05 \\ 
+  d & -0.23 & 0.08 & -3.02 & 0.00 \\ 
+  n1 & 0.00 & 0.00 & 1.74 & 0.08 \\ 
+  I(n1/n2) & -0.04 & 0.02 & -1.48 & 0.14 \\ 
+  density:d & 0.11 & 0.11 & 1.02 & 0.31 \\ 
+  dotplot:d & 0.20 & 0.11 & 1.86 & 0.06 \\ 
+  histogram:d & -0.01 & 0.11 & -0.09 & 0.93 \\ 
+  density:n1 & -0.00 & 0.00 & -2.15 & 0.03 \\ 
+  dotplot:n1 & -0.00 & 0.00 & -2.87 & 0.00 \\ 
+  histogram:n1 & -0.00 & 0.00 & -0.93 & 0.35 \\ 
+  density:I(n1/n2) & 0.04 & 0.03 & 1.18 & 0.24 \\ 
+  dotplot:I(n1/n2) & 0.02 & 0.03 & 0.65 & 0.52 \\ 
+  histogram:I(n1/n2) & 0.00 & 0.04 & 0.04 & 0.97 \\ 
+   \hline
+\end{tabular}
+\end{center}
+\caption{log time taken}
+
+\end{table}

### 2012-03-29 experiment.tex
working on intro, added/fixed references
second experiment

@@
    \centering
-   \includegraphics[width=3in]{turk4_samplesize_correctn.pdf}  
+   %\includegraphics[width=3in]{turk4_samplesize_correctn.pdf}  
    \caption{For each sample size for polar and Euclidian coordinates, the bars are colored according to the number of correct responses.
@@
 
-The effect of a reference line on accuracy and time was not significant for both types of charts.
+The effect of a reference line on accuracy and time was not statistically significant for both types of charts.
+When looking at plots of the data, however, the time spent on the charts seems to be higher when there is a reference line.This is true for both Euclidian and polar charts.
+Also visible in the data is a slight increase in accuracy with the presence of a reference line.
+These characteristics can be seen in figure\ref{reflines}.
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+   \includegraphics[width=2in]{turk4_refilne_time.pdf}  
+   \includegraphics[width=2in]{turk4_refline_correctn.pdf}  
+   \caption{The relationship between reference line and time spent (left) and the relationship between reference line and accuracy (right).
+Although not statistically significant, some patterns are visible.}
+   \label{reflines}
+\end{figure}

### 2012-03-29 experiment-two.tex
results for second experiment

@@
 
+\begin{figure*}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+   \includegraphics[width=\linewidth]{power-exp2} 
+   \caption{Overview of power predictions for the four different designs.
+The fully saturated thick lines show average predicted power for each of the designs facetted by size of the red group (top to bottom) and relative size of the blue group to the red group (left to right).
+}
+   \label{fig:power2}
+\end{figure*}
+
+Figure \ref{fig:power2} summarizes the modeling results as shown in table \ref{tbl:power2}.
+Generally we can see that the power of all four designs increases with an increase in effect size ($x$ axis), as well as an increase in sample size (top to bottom facetting).
+Depending on visualization setup, different designs have highest power: boxplots seem to be overall a good choice in visualizing a mean shift in a distribution.
+For small overall sample size dotplots (green dots and lines) show good power, but they  start to lose their advantage in favor of more summarized plots, such as density charts or histograms.
+Density charts are particularly suited for larger sample sizes.
 \begin{table}[ht]
@@
   \hline
- & Estimate & Std. Error & z value & Pr($>$$|$z$|$) \\ 
+ & Estimate & Std. Error & $z$ value & $p$-value \\ 
   \hline
-(Intercept) & -2.00 & 0.38 & -5.29 & 0.00 \\ 
-  density & -0.29 & 0.53 & -0.55 & 0.58 \\ 
-  dotplot & -0.72 & 0.52 & -1.39 & 0.17 \\ 
-  histogram & -0.29 & 0.53 & -0.54 & 0.59 \\ 
-  d & 1.78 & 0.34 & 5.22 & 0.00 \\ 
-  n1 & 0.02 & 0.00 & 9.77 & 0.00 \\ 
-  I(n1/n2) & -0.31 & 0.11 & -2.94 & 0.00 \\ 
-  density:d & -0.61 & 0.47 & -1.29 & 0.20 \\ 
-  dotplot:d & 0.04 & 0.46 & 0.08 & 0.94 \\ 
-  histogram:d & -0.29 & 0.48 & -0.60 & 0.55 \\ 
-  density:n1 & -0.00 & 0.00 & -0.40 & 0.69 \\ 
-  dotplot:n1 & -0.01 & 0.00 & -3.04 & 0.00 \\ 
-  histogram:n1 & -0.01 & 0.00 & -2.67 & 0.01 \\ 
-  density:I(n1/n2) & 0.09 & 0.15 & 0.59 & 0.55 \\ 
-  dotplot:I(n1/n2) & 0.27 & 0.15 & 1.81 & 0.07 \\ 
-  histogram:I(n1/n2) & 0.09 & 0.15 & 0.62 & 0.53 \\ 
+(Intercept) & -3.11 & 0.41 & -7.61 & 0.00 \\ 
+  density & 0.05 & 0.57 & 0.09 & 0.93 \\ 
+  dotplot & 0.46 & 0.57 & 0.81 & 0.42 \\ 
+  histogram & 0.06 & 0.58 & 0.11 & 0.91 \\ 
+  d & 1.77 & 0.34 & 5.18 & 0.00 \\ 
+  n1 & 0.02 & 0.00 & 9.73 & 0.00 \\ 
+  ratio & 0.84 & 0.33 & 2.56 & 0.01 \\ 
+  density:d & -0.60 & 0.47 & -1.27 & 0.20 \\ 
+  dotplot:d & 0.03 & 0.46 & 0.07 & 0.95 \\ 
+  histogram:d & -0.28 & 0.48 & -0.59 & 0.56 \\ 
+  density:n1 & -0.00 & 0.00 & -0.36 & 0.72 \\ 
+  dotplot:n1 & -0.01 & 0.00 & -3.00 & 0.00 \\ 
+  histogram:n1 & -0.01 & 0.00 & -2.64 & 0.01 \\ 
+  density:ratio & -0.28 & 0.47 & -0.59 & 0.56 \\ 
+  dotplot:ratio & -1.01 & 0.46 & -2.19 & 0.03 \\ 
+  histogram:ratio & -0.28 & 0.47 & -0.59 & 0.55 \\ 
    \hline
@@
 \end{center}
-\caption{correctness}
+\caption{Results from a Generalized Linear Mixed Effects Model of Power of a lineup given design and all two-way interactions with effect size $d$, group size $n_1$, and ratio to second to first group size.
+The model was fit based on 2513 lineup evaluations of 208 participants.}
+\label{tbl:power2}
 \end{table}

### 2012-03-29 experiment-two.tex
added discussion to second experiment

@@
 \subsection{Comparing Distributions}
-
-\begin{figure*} [hbtp]
+In a second experiment, we conducted a simulation study to investigate the power of four competing designs, as shown in figure\ref{fig:expii}, in assessing a mean shift between distributions.
+As covariates for this experiment, we considered the {\it size of the shift}, $d$, between the means of the distributions, with $d=0.4, 0.6, 0.8.
+1.0,$ and $1.2$, as well as the sample size.
+Since we are dealing with two groups of points, we varied both the size of the larger group, $n_1 = 15, 45$, and 135, as well as the relative size of the second group, with $n_2 = 1/3, 2/3$ and $3/3$ of $n_1$.
+For each of these combinations, we created three replicate data sets and four lineup charts, one for each of the competing designs.
+This resulted in 540 different lineup charts overall, which we sorted by $p$-value corresponding to the difference in means between the two groups in the simulated data.
+According to the $p$-value, we associated a `difficulty' level with lineups from 1 to 9 (easy to hard).
+Ten lineup plots were shown to each participant - one out of each of the difficulty levels and one additional reference chart with a particularly easy lineup  to allows for data quality checks.
+\begin{figure} [hbtp]
    \centering
-   \includegraphics[width=0.245\textwidth]{hist-conc} 
-   \includegraphics[width=0.245 \textwidth]{boxplot-conc} 
-   \includegraphics[width=0.245 \textwidth]{density-conc} 
-   \includegraphics[width=0.245 \textwidth]{dotplot-conc} 
+   \includegraphics[width=0.45\linewidth]{hist-conc} 
+   \includegraphics[width=0.45 \linewidth]{boxplot-conc} \\
+   \includegraphics[width=0.45 \linewidth]{density-conc} 
+   \includegraphics[width=0.45 \linewidth]{dotplot-conc} 
    \caption{Overview of all four competing designs for displaying differences in distributions.}
    \label{fig:expii}
-\end{figure*}
+\end{figure}
 
@@
 
+Since each participant provides results from nine lineups, we can use a generalized linear mixed effects model of power and account for individuals' abilities by including a subject-specific random intercept.
+We included all of the above described covariates  in the model, both as main effects and as two-way interaction with the design to assess their impact on the power of each of the designs.
 Figure \ref{fig:power2} summarizes the modeling results as shown in table \ref{tbl:power2}.
 Generally we can see that the power of all four designs increases with an increase in effect size ($x$ axis), as well as an increase in sample size (top to bottom facetting).
-Depending on visualization setup, different designs have highest power: boxplots seem to be overall a good choice in visualizing a mean shift in a distribution.
+Depending on visualization setup, different designs have highest power: boxplots  are an alround good choice in visualizing a mean shift in a distribution.
 For small overall sample size dotplots (green dots and lines) show good power, but they  start to lose their advantage in favor of more summarized plots, such as density charts or histograms.
-Density charts are particularly suited for larger sample sizes.
+Following the predicted power for density charts for the different ratios between the groups, we see that from left to right the density line does not benefit from additional information in the second group, unlike all of the other designs.
+Dotplots are the only design, in which each observation is shown individually.
+This increase in visual complexity might be the reason behind this phenomenon.
+
+Density charts profit the most from an increase in sample size - making them particularly good design choices for a larger sample sizes.
 \begin{table}[ht]

### 2012-03-29 intro.tex
typo in setup, and massaging intro

--- 2012-03-28 intro.tex
@@
 These approaches lack broad audience validity and
-objectiveness. Several papers indicate continuing issues arising with
-evaluating information visualizations
+objectiveness. Several papers indicate continuing issues arising with,
+and suggestions for improving, evaluation of information
+visualizations
 \citep{North:2006,Ellis:2006,Viegas:2006,Munzer:2009}.
@@
 
-Usually, we need simulation studies to be able to completely control for the signal strength in the data. 
+Here we suggest a new approach using recent research on visual
+inference \citep{buja:2009,wickham:2010}. In visual inference, real
+data plots are considered to be test statistics, and these are
+compared with plots of data generated from a null hypothesis using a
+lineup. A lineup is similar to the police lineup, and this is where
+the name comes from. The suspect (the plot of the real data) is placed
+randomly in a field of plots of null data, and an impartial viewer is
+asked to identify the plot that is the most different. If the viewer
+identifies the suspect it lends statistical signficance to the
+conclusion that the real data is not consistent with the null data.
+
+Lineups enable specific plot design evaluation of a specific
+visualization result. Different lineups are created, using the same
+data, same null plots and positions within the lineup, but different
+plot designs. Comparing how long and how accurately the viewers take
+to report the feature of interest will assess which design is better
+for the task. In addition, this approach is ideally coupled with
+simulation studies, where data is constructed to match and expand on
+the features of interest in the real data. For example, in a simple
+situation, where the discovered feature of interest is a difference
+between the centers of two distributions, simulated data would be
+constructed roughly matching the distribution of the real data that
+varies the distance between centers, and also spread and sample
+size. Lineups of these simulated data (embedded with simulated null
+data plots) are generated and evaluated along with the lineups of real
+data. The simulation study provides a backdrop for the real data,
+which enables the broad applicability to be studied and a guage for
+the signal strength of the pattern in the real data. It provides a way
+to objectively evaluate plot design.
+
+%Usually, we need simulation studies to be able to completely control for the signal strength in the data.
 For charts, this is not a very realistic scenario.
@@
 
-Lineup tests  \citep{buja:2009} allow us to directly use the data at hand to evaluate different competing designs by using permutations tests \cite{good:2011} of the original dataset.
+%Lineup \citep{buja:2009} allow us to directly use the data at hand to evaluate different competing designs by using permutations tests \cite{good:2011} of the original dataset.
 We can therefore get valid quantitative measures for comparing different designs without having to set up a simulation.
 
-When we are presenting a chart we  want to convince our audience of the presence of a particular feature or relationship in the data. 
+%When we are presenting a chart we  want to convince our audience of the presence of a particular feature or relationship in the data.
 In order to enable us to draw a conclusion about presence or absence of a feature in the data, we have, however, to assess what features show up based purely on marginal distributions rather than joint distributions.
 
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-   \includegraphics[height=0.32\linewidth]{rexp.pdf} 
-   \includegraphics[height=0.32\linewidth]{rnorm.pdf} 
-   \includegraphics[height=0.32\linewidth]{two-exp.pdf} 
-   \caption{Examples of charts with distinctive features but independent variables.
+%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+%   \centering
+%   \includegraphics[height=0.32\linewidth]{rexp.pdf} 
+%   \includegraphics[height=0.32\linewidth]{rnorm.pdf} 
+%   \includegraphics[height=0.32\linewidth]{two-exp.pdf} 
+%   \caption{Examples of charts with distinctive features but independent variables.
 From left to right: scatterplot of $X$ and $Y$ with independent $X, Y \sim Exp_{\lambda=1}$; scatterplot of two variables with mixed normal distribution $N(3,1)$ and $N(-3,1)$; boxplot of $X$ and $Y$ with $X \sim B_{n=100, p=0.1}$ and $Y \sim Exp_{\lambda=1}$.
 }
-   \label{random}
-\end{figure}
+%   \label{random}
+%\end{figure}
 
-Features in the marginal distributions, such as multi-modality, skewness  or outliers, might interfere with our intuition of what `randomness' might mean in a particular situation and lead us to wrong conclusions.
+%Features in the marginal distributions, such as multi-modality, skewness  or outliers, might interfere with our intuition of what `randomness' might mean in a particular situation and lead us to wrong conclusions.
 All of the plots in figure \ref{random} show distinctive features.

### 2012-03-29 conclusions.tex
finished re-working the intro

\section{Conclusions}

Could say something about adding a visual pleasure rating to the results, rate how attractive the display is on a scale of 1-5.

### 2012-03-29 experiment-two.tex
finished re-working the intro

@@
 % !TEX root = comparison.tex
-\subsection{Comparing Distributions}
+\subsection{Comparing Distributions}~\label{experimenttwo}
 In a second experiment, we conducted a simulation study to investigate the power of four competing designs, as shown in figure\ref{fig:expii}, in assessing a mean shift between distributions.

### 2012-03-29 intro.tex
finished re-working the intro

@@
 identifies the suspect it lends statistical signficance to the
-conclusion that the real data is not consistent with the null data.
+conclusion that the real data is not consistent with the null
+data. Figure \ref{lineup} has an example of a lineup. Overlaid density
+plots are used to display the distribution of two groups (blue, red)
+of data. We are interested to know whether the real data has a
+significant diffeerence between the centers of the two groups. Plot
+numbered 13 contains the real data. Did you pick that one? In this
+plot the blue group has a flat density whereas the density of red
+group is more peaked and concentrated to the left of the blue.
 
-Lineups enable specific plot design evaluation of a specific
-visualization result. Different lineups are created, using the same
-data, same null plots and positions within the lineup, but different
-plot designs. Comparing how long and how accurately the viewers take
-to report the feature of interest will assess which design is better
-for the task. In addition, this approach is ideally coupled with
-simulation studies, where data is constructed to match and expand on
-the features of interest in the real data. For example, in a simple
-situation, where the discovered feature of interest is a difference
-between the centers of two distributions, simulated data would be
-constructed roughly matching the distribution of the real data that
-varies the distance between centers, and also spread and sample
-size. Lineups of these simulated data (embedded with simulated null
-data plots) are generated and evaluated along with the lineups of real
-data. The simulation study provides a backdrop for the real data,
+\begin{figure}[htbp]
+   \centering
+   \includegraphics[height=0.95\linewidth]{density_1_45_15_20_13.png} 
+   \caption{Example lineup using density plots: one plot of real data
+     embedded with nineteen plots of null data. Which plot is
+     most different from the others? (See the text for the answer.)}
+   \label{lineup}
+\end{figure}
+
+The previous work \citet{buja:2009,wickham:2010} describes the use of
+lineups for evaluating discoveries made in exploratory analysis. Here
+we describe the use of lineups for evalating plot design for
+communicating findings visually. Different lineups are created, using
+the same data, same null plots and positions within the lineup, but
+different plot designs. Comparing how long and how accurately the
+viewers take to report the feature of interest will assess which
+design is better for the task. In addition, this approach is ideally
+coupled with simulation studies, where data is constructed to match
+and expand on the features of interest in the real data. For example,
+in a simple situation, where the discovered feature of interest is a
+difference between the centers of two distributions, simulated data
+would be constructed roughly matching the distribution of the real
+data that varies the distance between centers, and also spread and
+sample size. Lineups of these simulated data (embedded with simulated
+null data plots) are generated and evaluated along with the lineups of
+real data. The simulation study provides a backdrop for the real data,
 which enables the broad applicability to be studied and a guage for
@@
 
-Lineup plots have been shown as an effective tool \cite{wickham:2010} for enabling us to quantitively assess the strength of the relationship between $X$ and $Y$.
+%Lineup plots have been shown as an effective tool \cite{wickham:2010} for enabling us to quantitively assess the strength of the relationship between $X$ and $Y$.
 In a lineup plot the plot of the actual data is -- as in a police line-up -- inserted among decoy plots.
@@
 
-The strength of this evidence is determined by both the number of independent evaluations as well as the size of the lineup.
-Assume a lineup of size $m$ is evaluated by $n$ independent observers, out of which $x$ identify the data plot correctly, which we can use as an estimate for the probability to correctly identify the data plot:
-this leads us to an estimate for the probability of correctly identifying the plot of $x/n$, with 95\% confidence intervals  for $x/n$ given as $x/n \pm 1.96 \cdot \sqrt{x/n \cdot (1-x/n)}$ limited to the interval [0,1].
-Note, that the resulting interval is often too large to be useful.
-
-We can think of a lineup as $m-1$ head-to-head comparisons of the dataplot and $m-1$ nullplots, which the dataplot all has to `win', in order to be picked as the plot with the most distinct features.
-Assuming that observers identify the plot with the strongest signal, the probability that data plot is picked follows a Beta distribution for the maximum:
-this allows us to introduce the {\it signal strength} \cite{majumder:2011} of a lineup as:
+Lineups provide a way to quantify the significance of the finding
+\citet{majumder:2011}. The significance depends on the size ($m$) of
+the lineup ($m=20$ in Figure \ref{lineup}) and the number of
+independent observers ($n$). Let $x$ be the number of observers who
+correctly identify the plot of the real data. The probability that one
+observer correctly identifies the real data plot, in a size $m$
+lineup, by chance is $1/m$. To incorporate the results from multiple
+observers think about the lineup as $m-1$ head-to-head competitions of
+the data plot against $m-1$ null plots, which the data plot has to
+win, is picked as the plot with the most distinct features. Assume
+that the data plot is the one with the strongest signal, then the
+probability that the data plot is picked by $x$ observers follows a
+Beta distribution for the maximum \citet{****}:
 \[
@@
 \]
-Signal strength is limited to a range of values between 0 and 1. Higher values indicate stronger signal.
+We call it signal strength, and it is analgous to power in classical hypothesis testing.
+Signal strength will range between 0 and 1, with higher values indicating stronger signal, more distinctive feature in the data.
+
+*** The above paragraph needs more explanation, and a reference for the beta distribution.
+I don't think the Majumder paper really is the source for this result is it?
+It leads up to it, but this is the new work that is not submitted yet.
+
+*** Also need to refer to Niladri's ``Where's Waldo paper'' here too.
+
+%The strength of this evidence is determined by both the number of independent evaluations as well as the size of the lineup.
+Assume a lineup of size $m$ is evaluated by $n$ independent observers, out of which $x$ identify the data plot correctly, which we can use as an estimate for the probability to correctly identify the data plot: this leads us to an estimate for the probability of correctly identifying the plot of $x/n$, with 95\% confidence intervals  for $x/n$ given as $x/n \pm 1.96 \cdot \sqrt{x/n \cdot (1-x/n)}$ limited to the interval [0,1].
+Note, that the resulting interval is often too large to be useful.
+
+%We can think of a lineup as $m-1$ head-to-head comparisons of the dataplot and $m-1$ nullplots, which the dataplot all has to `win', in order to be picked as the plot with the most distinct features.
+Assuming that observers identify the plot with the strongest signal, the probability that data plot is picked follows a Beta distribution for the maximum: this allows us to introduce the {\it signal strength} \cite{majumder:2011} of a lineup as:
+
+%Signal strength is limited to a range of values between 0 and 1.
+Higher values indicate stronger signal.
 In the case of a parametric situation (i.e.
@@
 
-\section{Comparison of Designs}
+Signal strength is used in our new work to evaluate the effectiveness of different plot designs.
+This paper describes two examples where we had real data analysis problems and decisions to make in order to communicate the results.
+Section \ref{designs} describes the process for comparing designs.
+Sections \ref{experimentone} and \ref{experimenttwo} describe the two data analysis problems and the experiments conducted to evaluate the plots designs.
+Section \ref{results} presents the findings of the experiments.
+
+\section{Comparison of Designs}\label{designs}
 We are going to make use of the signal strength gained from multiple viewings of a lineup in order to evaluate competing designs as follows:

### 2012-03-29 experiment.tex
typo in setup, and massaging intro
finished re-working the intro
first experiment: discussion

--- 2012-03-27 experiment.tex
@@
-\section{Results and Discussion}
+\section{Results and Discussion}~\label{results}
 
-Euclidian coordinates performed better when speaking of efficiency in terms of accuracy and speed.
-76.92\% of the Euclidian charts shown resulted in the a correct identification of the real data while only 20.31\% of the polar charts did the same.
-A t-test comparing means for charts with polar coordinates versus charts with Euclidian coordinates resulted in a t statistic of 15.2013 with a p-value of 2.2e-16.
-Similarly, a t-test comparing means of time spent on charts with Euclidian coordinates versus charts with polar coordinates resulted in a p-value of 1.014e-11.
-The average of the log of time spent on each individual chart for charts with Euclidian coordinates is 3.53 minutes while it was 4.07 minutes for charts with polar coordinates.
-Euclidian coordinates resulted in a significantly more accurate identification of the real data set in significantly shorter time.
+Euclidean coordinates performed better when speaking of efficiency in terms of accuracy and speed.
+76.92\% of the Euclidean charts shown resulted in the a correct identification of the real data while only 20.31\% of the polar charts did the same.
+A t-test comparing means for charts with polar coordinates versus charts with Euclidean coordinates resulted in a t statistic of 15.2013 with a p-value of 2.2e-16.
+Similarly, a t-test comparing means of time spent on charts with Euclidean coordinates versus charts with polar coordinates resulted in a p-value of 1.014e-11.
+The average of the log of time spent on each individual chart for charts with Euclidean coordinates is 3.53 minutes while it was 4.07 minutes for charts with polar coordinates.
+Euclidean coordinates resulted in a significantly more accurate identification of the real data set in significantly shorter time.
 
@@
 These predicted values were obtained by using R package lme4 and a mixed effect model explaining correct responses by type of chart and the log of time spent while accounting for different levels of accuracy for each individual who participated in the experiment.
-The predicted values show that there is an overall higher accuracy level for Euclidian coordinates when compared to polar coordinates.
+The predicted values show that there is an overall higher accuracy level for Euclidean coordinates when compared to polar coordinates.
 For both chart types, as the time spent increases, the predicted accuracy decreases.
@@
    \centering
-   \includegraphics[width=3in]{turk4_samplesize_correctn.pdf}  
-   \caption{For each sample size for polar and Euclidian coordinates, the bars are colored according to the number of correct responses.
+   %\includegraphics[width=3in]{turk4_samplesize_correctn.pdf}  
+   \caption{For each sample size for polar and Euclidean coordinates, the bars are colored according to the number of correct responses.
 Darker colors indicate lower accuracy levels while lighter colors indicate higher accuracy levels.
@@
 
+Ideally, efficient detection of data characteristics will be possible at small sample sizes.
 The difficulty of the charts is largely determined by the sample size taken from the original data set.
 Samples of 24 \% generally result in higher accuracy as the extra information facilitates detection of a relationship.
-Figure \ref{accuracy_preds} shows the relationship between the number of correct responses relative to the total number of responses and how this relationship differs between Euclidian and polar coordinates.
-The bars in the Euclidian plot are overall a lighter color than they are in the polar plot.
+Figure \ref{accuracy_preds} shows the relationship between the number of correct responses relative to the total number of responses and how this relationship differs between Euclidean and polar coordinates.
+The bars in the Euclidean plot are overall a lighter color than they are in the polar plot.
 This indicates a higher level of accuracy at all sample sizes.
-For Euclidian coordinates, we can see a stair step pattern in the accuracy levels as sample size increases.
+For Euclidean coordinates, we can see a stair step pattern in the accuracy levels as sample size increases.
 Lower levels of accuracy, indicated by darker colors, become consistently less prevelant as sample size increases.
@@
 Sample sizes of 2\% and 4\% have very similar accuracy levels.
-After 6\% the stair step pattern becomes visible and similar to that of Euclidian coordinates.
-Sample size seems to have less of an effect on accuracy for polar coordinates than it does for Euclidian coordinates.
+After 6\% the stair step pattern becomes visible and similar to that of Euclidean coordinates.
+Sample size seems to have less of an effect on accuracy for polar coordinates than it does for Euclidean coordinates.
+Predicted values of the proportion of correct responses when explained by sample size, the type of chart, and the interaction between these two factors, show that there is a significant difference between the relationship between sample size and correct responses for polar and Euclidean coordinates.
+Indeed, the proportion of correct responses increases more rapidly as sample size increases for Euclidean coordinates than it does for polar coordinates.
+
+The effect of a reference line on accuracy and time was not statistically significant for both types of charts.
+When looking at plots of the data, however, the time spent on the charts seems to be higher when there is a reference line.This is true for both Euclidean and polar charts.
+Also visible in the data is a slight increase in accuracy with the presence of a reference line.
+These characteristics can be seen in figure\ref{reflines}.
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+   %\includegraphics[width=2in]{turk4_refilne_time.pdf}  
+   %\includegraphics[width=2in]{turk4_refline_correctn.pdf}  
+   \caption{The relationship between reference line and time spent (left) and the relationship between reference line and accuracy (right).
+Although not statistically significant, some patterns are visible.}
+   \label{reflines}
+\end{figure}

### 2012-03-29 comparison.tex
table formatting
added references, re-worked experiment 1

@@
 properties and then show results from two studies evaluating competing
-designs: both studies are designed to go to the threshold of our
+designs: both studies are designed to go to the limits of our
 perceptual abilities to highlight differences between designs. We use
@@
 \centering
-\includegraphics[height=1in]{polar}\hspace{0.05\linewidth}
-\includegraphics[height=1in]{eucline}
+\includegraphics[height=1.2in]{polar}\hspace{0.05\linewidth}
+\includegraphics[height=1.2in]{eucline}
 %\includegraphics[width=16cm]{CypressView.eps}
-\caption{teaser chart should include all different designs}
+\caption{Which of the three plots is different from the others? - Is this question easier to answer with the Polar Charts or the Barcharts?}
 }

### 2012-03-29 experiment-two.tex
rearranged paper

--- 2012-03-29 experiment-two.tex
@@
 % !TEX root = comparison.tex
+
 \subsection{Comparing Distributions}
-In a second experiment, we conducted a simulation study to investigate the power of four competing designs, as shown in figure\ref{fig:expii}, in assessing a mean shift between distributions.
-As covariates for this experiment, we considered the {\it size of the shift}, $d$, between the means of the distributions, with $d=0.4, 0.6, 0.8.
-1.0,$ and $1.2$, as well as the sample size.
-Since we are dealing with two groups of points, we varied both the size of the larger group, $n_1 = 15, 45$, and 135, as well as the relative size of the second group, with $n_2 = 1/3, 2/3$ and $3/3$ of $n_1$.
-For each of these combinations, we created three replicate data sets and four lineup charts, one for each of the competing designs.
-This resulted in 540 different lineup charts overall, which we sorted by $p$-value corresponding to the difference in means between the two groups in the simulated data.
-According to the $p$-value, we associated a `difficulty' level with lineups from 1 to 9 (easy to hard).
-Ten lineup plots were shown to each participant - one out of each of the difficulty levels and one additional reference chart with a particularly easy lineup  to allows for data quality checks.
-\begin{figure} [hbtp]
-   \centering
-   \includegraphics[width=0.45\linewidth]{hist-conc} 
-   \includegraphics[width=0.45 \linewidth]{boxplot-conc} \\
-   \includegraphics[width=0.45 \linewidth]{density-conc} 
-   \includegraphics[width=0.45 \linewidth]{dotplot-conc} 
-   \caption{Overview of all four competing designs for displaying differences in distributions.}
-   \label{fig:expii}
-\end{figure}
-
 \begin{figure*}[htbp] %  figure placement: here, top, bottom, or page

### 2012-03-29 experiment.tex
experiment discussion
rearranged paper

@@
 \section{Results and Discussion}~\label{results}
+
+\subsection{Wind Direction and Airport Efficiency}
+Figure \ref{fig:time} shows time taken until a decision was made in histograms, facetted by design and colored by correctness of answer.
+Because of the skewness of the distributions, times were log-transformed.
+Polar charts take on average more time to answer, but are answered with much lower precision.
+Figure \ref{fig:conf} shows two barcharts of confidence levels by task, again coloring is used for correctness of answers.
+Euclidean displays lead to a very bimodal distribution of confidence: participants are either very sure or not sure at all of their answer.
+Confidence levels in polar charts are distributed much more uniformly.
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+   \includegraphics[width=\linewidth]{time-answer} 
+   \caption{Histograms of time taken for answering lineups. On average the Euclidean design is answered faster and with higher power (percentage of correct answers). }
+   \label{fig:time}
+\end{figure}
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+   \includegraphics[width=\linewidth]{conf-answer} 
+   \caption{Barcharts of (subjective) confidence in answering correctly for each lineup.
+A strongly bimodal distribution is apparent, while confidence levels for polar charts are more uniform.
+There are only slight (and non significant) difference between confidence levels and correctness of results.}
+   \label{fig:conf}
+\end{figure}
+
+The perceptual tasks involved in decoding the designs fit into the general framework of the piechart versus barchart discussion.
+However, we are not dealing with assessing angles, which is known to be a perceptual harder task than assessments of heights or widths in bars, see e.g.
+\cite{cleveland:1984, robbins:2004, Kosslyn:2006, few:2009}.
+Instead, the perceptual tasks consists of comparisons along a common axis (in the barcharts) and a common origin (in the polar charts).
+The difference in designs is therefore based on how well we are able to judge deviations from a horizontal line compared to deviations from a circle.
+Based on this, the charts with added reference lines provide us with exactly the frame we compare with and should, therefore, be the `better' designs - either in speed or accuracy.
+
+%Most generally this experiment can be paralleled to the pie chart versus barchart discussion \cite{***}. 
+%The Euclidean coordinate plot is made up of bars where areas are colored proportionally according to percentages.
+The polar coordinate plot is made up of pie slices which are also colored proportionally according to the data.
+%Viewers of these charts will visually decode the areas of the bars and pie slices and make a judgement based on their decoding.
+According to  \citet[page=40]{kosslyn:2006}, area is usually perceived as a function of what the actual area is.
+This function is the area raised to an exponent of approximately 0.8 and then multiplied by a constant.
+However, when bars are parallel, the relative height of these bars is perceived very accurately.
+From this information, one could infer that the Euclidean coordinate plot may gain efficiency based on the height of the colored bars.
+
+For evaluating the results, we are using a generalized linear mixed effects modeling  approach \cite{pinheiro:2000} for all three of our response values, using the {\t R} package {\tt lme4} \cite{bates:2011}.
+For power predictions (cf.
+table \ref{tbl:correct}) a logistic regression was fitted for the competing designs, including covariates {\it sample size} (2, 4,6, 8, 10, and 24 percent of the data) and {\it shift} in wind direction (offset of 0, 90, 180, and 270 degrees), both as main effects and in two-way interaction effects with design to assess their effect on the power of designs.
+To adjust for individuals' ability, a random intercept was included in the model.
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+
+\begin{tabular}{cl}
+\phantom{\includegraphics[width=0.03\linewidth]{Polar_Line.pdf}} & \vspace{-0.035in} \multirow{10}{*}{\hspace{-0.25in}\includegraphics[width=0.9\linewidth]{turk4-designs.pdf}} \\
+\includegraphics[width=0.05\linewidth]{Polar_Line.pdf} \\
+\includegraphics[width=0.05\linewidth]{Polar_NoLine.pdf} \\
+\includegraphics[width=0.043\linewidth]{Euclidian_Line.pdf} \\
+\includegraphics[width=0.043\linewidth]{Euclidian_NoLine.pdf}\\
+\phantom{\includegraphics[width=0.015\linewidth]{Polar_Line.pdf}}\\
+\includegraphics[width=0.05\linewidth]{Polar_Line.pdf} \\
+\includegraphics[width=0.05\linewidth]{Polar_NoLine.pdf} \\
+\includegraphics[width=0.043\linewidth]{Euclidian_Line.pdf} \\
+\includegraphics[width=0.043\linewidth]{Euclidian_NoLine.pdf}\\
+  \end{tabular} 
+  \vspace{0.1in}
+   \caption{Power results for two competing designs: polar versus a Euclidean, with and without a reference line.
+Error bars indicate 95\% confidence intervals.
+Plots are facetted by percentage of data used.
+}
+   \label{fig:treatment}
+\end{figure}
+
+Overall, the results show huge differences in the power of designs between polar charts and Euclidean: Euclidean designs are significantly more powerful than polar charts, particularly so with small sample sizes.
+The reference line has surprisingly little influence, but it helps more for  polar charts than for barcharts,
+An increase in sample size has positive impact on power (see figure \ref{fig:treatment} for an overview).
+Polar charts need a much bigger sample size to see an increase in power -- only at about 24\% of the original data do we see about the same power as for Euclidean charts of a sample size of 2\%.
+The  changes in offset are significant -- interestingly, borderline behavior (90 and 270 degrees) does not show a difference between polar and Euclidean charts, whereas an inversion of the wave pattern (first up, then down), does show a difference.
+The Power of Barcharts suffers significantly  from this inversion whereas power of polar charts is unaffected.
+This is an unexpected finding, but is consistent throughout different lineups in the data.
+Figure \ref{fig:power} summarizes the results from the model: power predictions ($y$ axis) are shown by sample size ($x$ axis).
+The thick lines show average power by design for different shifts in wind direction.
+The thin lines represent power for individuals.
+What can be seen is the different impact of the offset by design: while  an offset of 270 degrees (the 'mountain' pattern) has the highest power in Euclidean charts, it comes out worst in polar charts.
+
+\begin{figure*}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+   \includegraphics[width=\linewidth]{predict-power} 
+   \caption{Predicted Power of designs.
+The thin lines and the points on $y$ axis show  variability due to individuals' abilities.
+The saturated lines show average predicted power for each of the designs.}
+   \label{fig:power}
+\end{figure*}
+
+Time taken to answer was log transformed before the modeling process to de-emphasize the impact of very large values (up to 500 sec).
+The findings are consistent with correctness - time taken shows big differences between polar and Euclidean charts.
+The reference lines seem to increase the evaluation time, but not significantly.
+An increase in sample size decreases evaluation time, The inversion of the wave pattern at 180 degrees leads to a significant increase in time for Euclidean charts, but not for polar charts (at least not significantly).
+
+ Confidence levels are measured on a five point scale --- they are subjective assessments by the participant `how certain are you'.
+We model confidence using the same model structure as before, i.e.
+using sample size and offset as covariates and including up to two-way interactions,
+While there is a difference between the designs - participants reported higher confidence in dealing with the barcharts -- this is only  a trend (i.e.
+not significant at 5\% but below 10\%).
+The only significant effect on reported confidence level is the use of reference lines in polar coordinates: participants report an increase in confidence (0.31 $\pm$ 0.13, $p$ value = 0.01) when using the reference line.
+This, however, does not translate to a significant increase in accuracy of results, as we have seen before.
+
+\begin{table}[ht]
+\begin{center}
+\resizebox{\linewidth}{!} {
+%\rowcolors{2}{white}{lightgray}
+\begin{tabular}{rrrrrrl}
+  \hline
+& & Estimate & Error & $z$-value & $p$-value &\\ 
+  \hline
+\bf design & euclid & -0.08 & 0.39 & -0.21 & 0.84 &\\ 
+&polar & -1.98 & 0.32 & -6.13 & 0.00 & ***\\ [3pt]
+\multicolumn{2}{l}{\bf main effects} &&&&&\\
+& reference line  & -0.14 & 0.26 & -0.53 & 0.59 & \\ [1pt]
+&  sample size & 0.27 & 0.04 & 6.31 & 0.00 & ***\\[1pt]
+ &offset:  90 degrees& -0.43 & 0.37 & -1.18 & 0.24 &\\ 
+  & 180 degrees& -0.89 & 0.35 & -2.51 & 0.01 & **\\ 
+  & 270 degrees& 0.21 & 0.38 & 0.55 & 0.58 &\\ [3pt]
+\multicolumn{2}{l}{\bf interactions} &&&&&\\
+&  polar:line & 0.51 & 0.35 & 1.44 & 0.15 &\\ [1pt]
+&    polar:sample size & -0.23 & 0.05 & -5.02 & 0.00 & ***\\[1pt]
+&    polar:offset 90 & 0.64 & 0.49 & 1.30 & 0.20 \\ 
+&  polar:offset 180 & 0.91 & 0.47 & 1.92 & 0.05 & .\\ 
+&    polar:offset 270 & -0.73 & 0.54 & -1.35 & 0.18 &\\
+   \hline
+\\[-5pt]
+   \multicolumn{5}{l}{Signif. codes:  0 `***' 0.001 `**' 0.01 `*' 0.05 `.' 0.1}
+\end{tabular}
+}
+\end{center}
+\caption{\label{tbl:correct} Output of a generalized linear mixed effects model for power of lineups (i.e.
+probability of identifying the data plot) for comparing designs.
+Included are two-way interactions with sample size and shifts in wind direction (offset).
+Euclidean designs without a reference line at offset 0 are used as
+ Results are based on  976 lineup evaluations by 115 participants. }
+\end{table}
+
+\begin{table}[ht]
+\begin{center}
+\resizebox{\linewidth}{!} {
+%\rowcolors{2}{white}{lightgray}
+\begin{tabular}{lrrrrrl}
+  \hline
+& & Estimate & Error & $t$-value & \multicolumn{2}{l}{approx $p$-value} \\ 
+  \hline
+\bf design & Intercept & 3.58 & 0.09 & 41.34 & 0.00 & ***\\ 
+&  polar & 0.22 & 0.10 & 2.17 & 0.03 & * \\ [2pt]
+\multicolumn{2}{l}{\bf covariates}\\
+&reference line & 0.08 & 0.06 & 1.50 & 0.13 \\ [1pt]
+ & sample size & -0.03 & 0.00 & -8.73 & 0.00 & ***\\ [1pt]
+&  offset: 90 & -0.06 & 0.08 & -0.72 & 0.47 \\ 
+& 180 & 0.16 & 0.08 & 2.07 & 0.04 & *\\ 
+&  270 & -0.04 & 0.07 & -0.54 & 0.59 \\ [2pt]
+\multicolumn{2}{l}{\bf interactions}\\
+&  polar:line & -0.03 & 0.08 & -0.40 & 0.69 \\ [1pt]
+&    polar:sample size & 0.03 & 0.01 & 6.25 & 0.00 & ***\\ [1pt]
+&    polar:offset 90 & 0.14 & 0.11 & 1.23 & 0.22 \\ 
+&    polar:offset 180 & -0.17 & 0.11 & -1.55 & 0.12 \\ 
+&    polar:offset 270 & 0.17 & 0.11 & 1.50 & 0.13 \\ 
+   \hline
+\\[-5pt]
+   \multicolumn{5}{l}{Signif. codes:  0 `***' 0.001 `**' 0.01 `*' 0.05 `.' 0.1}
+\end{tabular}}
+\end{center}
+\caption{\label{tbl:time} Model output for linear mixed effects model of (log) time taken, barchart design is baseline. }
+\end{table}
 
@@
    \centering
-   %\includegraphics[width=2in]{turk4_refilne_time.pdf}  
-   %\includegraphics[width=2in]{turk4_refline_correctn.pdf}  
+   %\includegraphics[width=1.5in]{turk4_refilne_time.pdf}  
+   %\includegraphics[width=1.5in]{turk4_refline_correctn.pdf}  
    \caption{The relationship between reference line and time spent (left) and the relationship between reference line and accuracy (right).

### 2012-03-30 experiment-two.tex
discussion of second experiment: time and confidence

@@
 The fully saturated thick lines show average predicted power for each of the designs facetted by size of the red group (top to bottom) and relative size of the blue group to the red group (left to right).
+Thin lines represent variability due to subject-specific abilities.
 }
@@
 Figure \ref{fig:power2} summarizes the modeling results as shown in table \ref{tbl:power2}.
-Generally we can see that the power of all four designs increases with an increase in effect size ($x$ axis), as well as an increase in sample size (top to bottom facetting).
-Depending on visualization setup, different designs have highest power: boxplots  are an alround good choice in visualizing a mean shift in a distribution.
-For small overall sample size dotplots (green dots and lines) show good power, but they  start to lose their advantage in favor of more summarized plots, such as density charts or histograms.
-Following the predicted power for density charts for the different ratios between the groups, we see that from left to right the density line does not benefit from additional information in the second group, unlike all of the other designs.
-Dotplots are the only design, in which each observation is shown individually.
-This increase in visual complexity might be the reason behind this phenomenon.
+Generally we can see that the power of all four designs increases with an increase in effect size $d$ (on the $x$ axis), as well as an increasing sample size $n_1$ (top to bottom facetting).
+Generally, an increase of the second group also is beneficial to power.
+Unlike all of the other designs,  dotplots actually lose power significantly as the second group increases.
+The reason behind this phenomenon might be visual complexity: dotplots are the only design that show individuals.
+With an increase in group size, visual complexity increases with it, which could hamper the observer in evaluating shifts.
 
-Density charts profit the most from an increase in sample size - making them particularly good design choices for a larger sample sizes.
+The two highly summarized designs benefit the most from an increase in sample size: both density charts and boxplots profit significantly more than histograms and dotplots.
+This might, again, be a hint towards a trade-off between visual complexity and amount of information shown as basis for decisions.
 \begin{table}[ht]
 \begin{center}
-\begin{tabular}{rrrrr}
+\resizebox{\linewidth}{!} {
+\begin{tabular}{lrrrrrl}
   \hline
- & Estimate & Std. Error & $z$ value & $p$-value \\ 
+& & Estimate & Error & $t$-value & \multicolumn{2}{l}{approx $p$-value} \\   \hline
+\\[-8pt]
+\bf design& Intercept & -3.11 & 0.41 & -7.61 & 0.00 & ***\\ 
+&  density & 0.05 & 0.57 & 0.09 & 0.93 \\ 
+ & dotplot & 0.46 & 0.57 & 0.81 & 0.42 \\ 
+  &histogram & 0.06 & 0.58 & 0.11 & 0.91 \\ [2pt]
+\multicolumn{2}{l}{\bf covariates}\\
+&  d & 1.77 & 0.34 & 5.18 & 0.00 & ***\\ [1pt]
+&   n1$_{100}$ & 1.84 & 0.19 & 9.73 & 0.00 & *** \\ [1pt]
+&  ratio & 0.84 & 0.33 & 2.56 & 0.01 & ** \\ [2pt]
+\multicolumn{2}{l}{\bf interactions}\\
+&    density:d & -0.60 & 0.47 & -1.27 & 0.20 \\ 
+&    dotplot:d & 0.03 & 0.46 & 0.07 & 0.95 \\ 
+&    histogram:d & -0.28 & 0.48 & -0.59 & 0.56 \\ [1pt]
+&    density: n1$_{100}$ & -0.10 & 0.26 & -0.36 & 0.72 \\ 
+&    dotplot: n1$_{100}$ & -0.77 & 0.26 & -3.00 & 0.00 & *** \\ 
+&    histogram: n1$_{100}$ & -0.69 & 0.26 & -2.64 & 0.01 & ** \\ [1pt]
+&    density:ratio & -0.28 & 0.47 & -0.59 & 0.56 \\ 
+&    dotplot:ratio & -1.01 & 0.46 & -2.19 & 0.03 & * \\ 
+&histogram:ratio & -0.28 & 0.47 & -0.59 & 0.55 \\ 
    \hline
-(Intercept) & -3.11 & 0.41 & -7.61 & 0.00 \\ 
-  density & 0.05 & 0.57 & 0.09 & 0.93 \\ 
-  dotplot & 0.46 & 0.57 & 0.81 & 0.42 \\ 
-  histogram & 0.06 & 0.58 & 0.11 & 0.91 \\ 
-  d & 1.77 & 0.34 & 5.18 & 0.00 \\ 
-  n1 & 0.02 & 0.00 & 9.73 & 0.00 \\ 
-  ratio & 0.84 & 0.33 & 2.56 & 0.01 \\ 
-  density:d & -0.60 & 0.47 & -1.27 & 0.20 \\ 
-  dotplot:d & 0.03 & 0.46 & 0.07 & 0.95 \\ 
-  histogram:d & -0.28 & 0.48 & -0.59 & 0.56 \\ 
-  density:n1 & -0.00 & 0.00 & -0.36 & 0.72 \\ 
-  dotplot:n1 & -0.01 & 0.00 & -3.00 & 0.00 \\ 
-  histogram:n1 & -0.01 & 0.00 & -2.64 & 0.01 \\ 
-  density:ratio & -0.28 & 0.47 & -0.59 & 0.56 \\ 
-  dotplot:ratio & -1.01 & 0.46 & -2.19 & 0.03 \\ 
-  histogram:ratio & -0.28 & 0.47 & -0.59 & 0.55 \\ 
-   \hline
-\end{tabular}
+\\[-5pt]
+   \multicolumn{5}{l}{Signif. codes:  0 `***' 0.001 `**' 0.01 `*' 0.05 `.' 0.1}
+\end{tabular}}
 \end{center}
 \caption{Results from a Generalized Linear Mixed Effects Model of Power of a lineup given design and all two-way interactions with effect size $d$, group size $n_1$, and ratio to second to first group size.
-The model was fit based on 2513 lineup evaluations of 208 participants.}
+Reference group for all effects are boxplot designs.
+The model was fit based on 2513 lineup evaluations of 208 participants.
+Note that estimate and error for all terms involving $n_1$ are multiplied by a factor of 100.}
 \label{tbl:power2}
@@
 
+Table \ref{tbl:speed2} gives an overview of results from a linear model for (log) evaluation time of lineups.
+Subject-specific speed (due to spatial ability) are accounted for  including a random intercept.
+Structurally, the model is the same as before, i.e.
+we regard effect size $d$, size $n_1$ of the first group, and the ratio $n_2/n_1$ between group sizes both as main effects and in two-way interactions with the four competing designs.
+To make estimate sizes comparable, we have again multiplied all estimates and standard errors involving the group size $n_1$ by 100.
+This means that we need to interpret those  estimates as average change if the group size is increased by 100 elements.
+The results of this model are similar to the previous: evaluation speeds needed for the four different designs is very close, but  histograms take on average a significantly longer time ($e^{0.23} = 1.26$ seconds) to evaluate than lineups of boxplots.
+Where an increase of sample size led to an increase of accuracy before, we now see that it increases time for evaluation of boxplot slightly, but not significantly so.
+What is interesting is that evaluation time decreases significantly with an increase in sample size, particularly for dotplots, but also for density charts.
+This is counter-intuitive to our previous argument on dotplots being affected by visual complexity the most.
+It seems, that observers of dotplot lineups come to a conclusion faster as sample size increases, albeit more often to  the wrong conclusion.
+Rather than visual complexity, this might be indicative that over-plotting is problematic in dotplots.
+While we tried to accommodate for over-plotting by making use of alpha-blending and jittering, this might not have been sufficient for the large sample size.
+
 \begin{table}[ht]
 \begin{center}
-\begin{tabular}{rrrrr}
+\resizebox{\linewidth}{!} {
+\begin{tabular}{lrrrrrl}
   \hline
- & Estimate & Std..Error & t.value & pval \\ 
+& & Estimate & Error & $t$-value & \multicolumn{2}{l}{approx $p$-value} \\   \hline
+\\[-8pt]
+\bf design& Intercept & 3.85 & 0.09 & 42.21 & 0.00 & ***\\ 
+&  density & 0.01 & 0.12 & 0.07 & 0.94 \\ 
+&  dotplot & 0.03 & 0.12 & 0.26 & 0.80 \\ 
+&  histogram & 0.23 & 0.12 & 1.97 & 0.05 & . \\ [2pt]
+\multicolumn{2}{l}{\bf covariates}\\
+&      d & -0.23 & 0.08 & -3.02 & 0.00 & ***\\ [1pt]
+&    n1$_{100}$ & 0.07 & 0.04 & 1.70 & 0.09 & .\\ [1pt]
+&  ratio & -0.04 & 0.02 & -1.48 & 0.14  \\ [2pt]
+\multicolumn{2}{l}{\bf interactions}\\
+&  density:d & 0.11 & 0.11 & 1.02 & 0.31 \\ 
+&  dotplot:d & 0.20 & 0.11 & 1.86 & 0.06 & .  \\ 
+&  histogram:d & -0.01 & 0.11 & -0.09 & 0.93 \\ [1pt]
+&  density:n1$_{100}$ & -0.12 & 0.06 & -2.11 & 0.03 & *\\ 
+&  dotplot:n1$_{100}$ & -0.16 & 0.06 & -2.83 & 0.00 & ***\\ 
+&  histogram:n1$_{100}$ & -0.05 & 0.06 & -0.93 & 0.35 \\ [1pt]
+&  density:ratio & 0.04 & 0.03 & 1.18 & 0.24 \\ 
+&  dotplot:ratio & 0.02 & 0.03 & 0.65 & 0.52 \\ 
+&  histogram:ratio & 0.00 & 0.04 & 0.04 & 0.97 \\ 
    \hline
-(Intercept) & 3.85 & 0.09 & 42.21 & 0.00 \\ 
-  density & 0.01 & 0.12 & 0.07 & 0.94 \\ 
-  dotplot & 0.03 & 0.12 & 0.26 & 0.80 \\ 
-  histogram & 0.23 & 0.12 & 1.97 & 0.05 \\ 
-  d & -0.23 & 0.08 & -3.02 & 0.00 \\ 
-  n1 & 0.00 & 0.00 & 1.74 & 0.08 \\ 
-  I(n1/n2) & -0.04 & 0.02 & -1.48 & 0.14 \\ 
-  density:d & 0.11 & 0.11 & 1.02 & 0.31 \\ 
-  dotplot:d & 0.20 & 0.11 & 1.86 & 0.06 \\ 
-  histogram:d & -0.01 & 0.11 & -0.09 & 0.93 \\ 
-  density:n1 & -0.00 & 0.00 & -2.15 & 0.03 \\ 
-  dotplot:n1 & -0.00 & 0.00 & -2.87 & 0.00 \\ 
-  histogram:n1 & -0.00 & 0.00 & -0.93 & 0.35 \\ 
-  density:I(n1/n2) & 0.04 & 0.03 & 1.18 & 0.24 \\ 
-  dotplot:I(n1/n2) & 0.02 & 0.03 & 0.65 & 0.52 \\ 
-  histogram:I(n1/n2) & 0.00 & 0.04 & 0.04 & 0.97 \\ 
-   \hline
-\end{tabular}
+\\[-5pt]
+   \multicolumn{5}{l}{Signif. codes:  0 `***' 0.001 `**' 0.01 `*' 0.05 `.' 0.1}
+\end{tabular}}
 \end{center}
-\caption{log time taken}
+\caption{\label{tbl:speed2}Results from a Generalized Linear Mixed Effects Model of (log) time to answer a lineup given design and all two-way interactions with effect size $d$, group size $n_1$, and ratio to second to first group size.
+Estimate and error for all terms involving $n_1$ are multiplied by a factor of 100 for comparability.
+Reference group for all effects are boxplot designs. }
+\end{table}
 
-\end{table}
+Our third response variable is again the subjective assessment of confidence level.
+On average participants reported a confidence in their results of  3.41 (on a five point scale with 1=lowest and 5=highest), with a standard deviation of 0.149.
+Modeling confidence levels with a linear mixed effects model with the same structure as before resulted in only one significant effect beyond the intercept: participants report significantly higher confidence for increased number of  points in dotplot lineups.
+On average an increase of 100 points leads to an increase in confidence of 0.22 (with a standard error of 0.082).
+This finding is significant with a $p$-value of 0.007.
+Together with the previous results, we can tell that something interesting happens in the perception of dotplots as sample sizes increase. 
+*** Follow-up experiments are planned. *** do we?
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+   \includegraphics[width=\linewidth]{conf-exp2} 
+   \caption{example caption}
+   \label{fig:example}
+\end{figure}

### 2012-03-30 experiment.tex
discussion of second experiment: time and confidence

@@
+% !TEX root = comparison.tex
 \section{Results and Discussion}~\label{results}
@@
 
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+The perceptual tasks involved in decoding the designs fit into the general framework of the piechart versus barchart discussion.
+However, we are not dealing with assessing angles, which is known to be a perceptual harder task than assessments of heights or widths in bars, see e.g.
+\cite{cleveland:1984, robbins:2004, Kosslyn:2006, few:2009}.
+Instead, the perceptual tasks consists of comparisons along a common axis (in the barcharts) and a common origin (in the polar charts).
+The difference in designs is therefore based on how well we are able to judge deviations from a horizontal line compared to deviations from a circle.
+Based on this, the charts with added reference lines provide us with exactly the frame we compare with and should, therefore, be the `better' designs - either in speed or accuracy.
+
+\begin{figure}[hbtp] %  figure placement: here, top, bottom, or page
    \centering
@@
 
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+\begin{figure}[hbtp] %  figure placement: here, top, bottom, or page
    \centering
@@
 \end{figure}
-
-The perceptual tasks involved in decoding the designs fit into the general framework of the piechart versus barchart discussion.
-However, we are not dealing with assessing angles, which is known to be a perceptual harder task than assessments of heights or widths in bars, see e.g.
-\cite{cleveland:1984, robbins:2004, Kosslyn:2006, few:2009}.
-Instead, the perceptual tasks consists of comparisons along a common axis (in the barcharts) and a common origin (in the polar charts).
-The difference in designs is therefore based on how well we are able to judge deviations from a horizontal line compared to deviations from a circle.
-Based on this, the charts with added reference lines provide us with exactly the frame we compare with and should, therefore, be the `better' designs - either in speed or accuracy.
 
@@
 Overall, the results show huge differences in the power of designs between polar charts and Euclidean: Euclidean designs are significantly more powerful than polar charts, particularly so with small sample sizes.
-The reference line has surprisingly little influence, but it helps more for  polar charts than for barcharts,
+76.92\% of the Euclidean lineups shown resulted in the a correct identification of the real data while only 20.31\% of the polar charts were correct.
+ The reference line has surprisingly little influence, but it helps more for  polar charts than for Euclidean charts,
 An increase in sample size has positive impact on power (see figure \ref{fig:treatment} for an overview).
@@
 The  changes in offset are significant -- interestingly, borderline behavior (90 and 270 degrees) does not show a difference between polar and Euclidean charts, whereas an inversion of the wave pattern (first up, then down), does show a difference.
-The Power of Barcharts suffers significantly  from this inversion whereas power of polar charts is unaffected.
+The power of Euclidean lineups suffers significantly  from this inversion whereas power of polar charts is unaffected.
 This is an unexpected finding, but is consistent throughout different lineups in the data.
@@
 
-\begin{figure*}[htbp] %  figure placement: here, top, bottom, or page
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
    \centering
@@
    \label{fig:power}
-\end{figure*}
+\end{figure}
 
@@
 The findings are consistent with correctness - time taken shows big differences between polar and Euclidean charts.
+The average of the amount of time spent on a Euclidean lineup is 3.53 minutes while it was 4.07 minutes for a polar lineup.
 The reference lines seem to increase the evaluation time, but not significantly.
@@
 using sample size and offset as covariates and including up to two-way interactions,
-While there is a difference between the designs - participants reported higher confidence in dealing with the barcharts -- this is only  a trend (i.e.
+While there is a difference between the designs - participants reported higher confidence in dealing with the Euclidean lineups -- this is only  a trend (i.e.
 not significant at 5\% but below 10\%).
@@
 This, however, does not translate to a significant increase in accuracy of results, as we have seen before.
+
+ Euclidean coordinates resulted in a significantly more accurate identification of the real data set in significantly shorter time.
 
@@
   \hline
-& & Estimate & Error & $t$-value & \multicolumn{2}{l}{approx $p$-value} \\ 
-  \hline
+& & Estimate & Error & $t$-value & \multicolumn{2}{l}{approx $p$-value} \\   \hline
 \bf design & Intercept & 3.58 & 0.09 & 41.34 & 0.00 & ***\\ 
@@
 
-Euclidean coordinates performed better when speaking of efficiency in terms of accuracy and speed.
+%Euclidean coordinates performed better when speaking of efficiency in terms of accuracy and speed.
 76.92\% of the Euclidean charts shown resulted in the a correct identification of the real data while only 20.31\% of the polar charts did the same.
@@
 
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-   \includegraphics[width=3in]{turk4_time_perc_preds.pdf}  
-   \caption{Values predicted from XXXX how to say itXXXX of accuracy for log of time taken and test parameter.}
-   \label{accuracy_preds}
-\end{figure}
-
-Figure \ref{accuracy_preds} shows predicted accuracy levels.
+%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+%   \centering
+%   \includegraphics[width=3in]{turk4_time_perc_preds.pdf}  
+%   \caption{Values predicted from XXXX how to say itXXXX of accuracy for log of time taken and test parameter.}
+%   \label{accuracy_preds}
+%\end{figure}
+%
+%Figure \ref{accuracy_preds} shows predicted accuracy levels.
 These predicted values were obtained by using R package lme4 and a mixed effect model explaining correct responses by type of chart and the log of time spent while accounting for different levels of accuracy for each individual who participated in the experiment.
@@
 For both chart types, as the time spent increases, the predicted accuracy decreases.
-
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-   %\includegraphics[width=3in]{turk4_samplesize_correctn.pdf}  
-   \caption{For each sample size for polar and Euclidean coordinates, the bars are colored according to the number of correct responses.
+%
+%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+%   \centering
+%   %\includegraphics[width=3in]{turk4_samplesize_correctn.pdf}  
+%   \caption{For each sample size for polar and Euclidean coordinates, the bars are colored according to the number of correct responses.
 Darker colors indicate lower accuracy levels while lighter colors indicate higher accuracy levels.
 }
-   \label{samp_size_acc}
-\end{figure}
+%   \label{samp_size_acc}
+%\end{figure}
 
-Ideally, efficient detection of data characteristics will be possible at small sample sizes.
+%Ideally, efficient detection of data characteristics will be possible at small sample sizes.
 The difficulty of the charts is largely determined by the sample size taken from the original data set.
@@
 For Euclidean coordinates, we can see a stair step pattern in the accuracy levels as sample size increases.
-Lower levels of accuracy, indicated by darker colors, become consistently less prevelant as sample size increases.
+Lower levels of accuracy, indicated by darker colors, become consistently less prevalent as sample size increases.
 The relationship between sample size and accuracy level is not as simple for polar coordinates.
@@
 Indeed, the proportion of correct responses increases more rapidly as sample size increases for Euclidean coordinates than it does for polar coordinates.
-
-The effect of a reference line on accuracy and time was not statistically significant for both types of charts.
+%
+%The effect of a reference line on accuracy and time was not statistically significant for both types of charts.
 When looking at plots of the data, however, the time spent on the charts seems to be higher when there is a reference line.This is true for both Euclidean and polar charts.
@@
 These characteristics can be seen in figure\ref{reflines}.
-
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-   %\includegraphics[width=1.5in]{turk4_refilne_time.pdf}  
-   %\includegraphics[width=1.5in]{turk4_refline_correctn.pdf}  
-   \caption{The relationship between reference line and time spent (left) and the relationship between reference line and accuracy (right).
+%
+%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+%   \centering
+%   %\includegraphics[width=1.5in]{turk4_refilne_time.pdf}  
+%   %\includegraphics[width=1.5in]{turk4_refline_correctn.pdf}  
+%   \caption{The relationship between reference line and time spent (left) and the relationship between reference line and accuracy (right).
 Although not statistically significant, some patterns are visible.}
-   \label{reflines}
-\end{figure}
+%   \label{reflines}
+%\end{figure}

### 2012-03-30 experiment-two.tex
some more on confidence levels in experiment 2

@@
 This might, again, be a hint towards a trade-off between visual complexity and amount of information shown as basis for decisions.
-\begin{table}[ht]
-\begin{center}
-\resizebox{\linewidth}{!} {
-\begin{tabular}{lrrrrrl}
-  \hline
-& & Estimate & Error & $t$-value & \multicolumn{2}{l}{approx $p$-value} \\   \hline
-\\[-8pt]
-\bf design& Intercept & -3.11 & 0.41 & -7.61 & 0.00 & ***\\ 
-&  density & 0.05 & 0.57 & 0.09 & 0.93 \\ 
- & dotplot & 0.46 & 0.57 & 0.81 & 0.42 \\ 
-  &histogram & 0.06 & 0.58 & 0.11 & 0.91 \\ [2pt]
-\multicolumn{2}{l}{\bf covariates}\\
-&  d & 1.77 & 0.34 & 5.18 & 0.00 & ***\\ [1pt]
-&   n1$_{100}$ & 1.84 & 0.19 & 9.73 & 0.00 & *** \\ [1pt]
-&  ratio & 0.84 & 0.33 & 2.56 & 0.01 & ** \\ [2pt]
-\multicolumn{2}{l}{\bf interactions}\\
-&    density:d & -0.60 & 0.47 & -1.27 & 0.20 \\ 
-&    dotplot:d & 0.03 & 0.46 & 0.07 & 0.95 \\ 
-&    histogram:d & -0.28 & 0.48 & -0.59 & 0.56 \\ [1pt]
-&    density: n1$_{100}$ & -0.10 & 0.26 & -0.36 & 0.72 \\ 
-&    dotplot: n1$_{100}$ & -0.77 & 0.26 & -3.00 & 0.00 & *** \\ 
-&    histogram: n1$_{100}$ & -0.69 & 0.26 & -2.64 & 0.01 & ** \\ [1pt]
-&    density:ratio & -0.28 & 0.47 & -0.59 & 0.56 \\ 
-&    dotplot:ratio & -1.01 & 0.46 & -2.19 & 0.03 & * \\ 
-&histogram:ratio & -0.28 & 0.47 & -0.59 & 0.55 \\ 
-   \hline
-\\[-5pt]
-   \multicolumn{5}{l}{Signif. codes:  0 `***' 0.001 `**' 0.01 `*' 0.05 `.' 0.1}
-\end{tabular}}
-\end{center}
-\caption{Results from a Generalized Linear Mixed Effects Model of Power of a lineup given design and all two-way interactions with effect size $d$, group size $n_1$, and ratio to second to first group size.
-Reference group for all effects are boxplot designs.
-The model was fit based on 2513 lineup evaluations of 208 participants.
-Note that estimate and error for all terms involving $n_1$ are multiplied by a factor of 100.}
-\label{tbl:power2}
-\end{table}
 
@@
 This finding is significant with a $p$-value of 0.007.
-Together with the previous results, we can tell that something interesting happens in the perception of dotplots as sample sizes increase. 
-*** Follow-up experiments are planned. *** do we?
+Together with the previous results, we can tell that something worth of further investigation happens in the perception of dotplots as sample sizes increase.
 
@@
    \includegraphics[width=\linewidth]{conf-exp2} 
-   \caption{example caption}
-   \label{fig:example}
+   \caption{Barcharts of marginal distributions of subjective confidence assessment by design. The distributions display prominent differences.}
+   \label{fig:conf-margins}
 \end{figure}
+
+Figure \ref{fig:conf-margins} shows barcharts of confidence assessments by design.
+It is obvious that the distribution of confidence is very different from one design to the next.
+Boxplots and Dotplots show a similar pattern with an emphasis on high confidence levels, whereas histograms and density lineups have modes at a confidence level of three.
+One reason behind these differences in confidence distributions might be familiarity with the design - neither density charts nor histograms are particularly frequent displays for a general audience.

### 2012-03-30 experiment-two.tex
table back in

@@
 With an increase in group size, visual complexity increases with it, which could hamper the observer in evaluating shifts.
+
+\begin{table}[ht]
+\begin{center}
+\resizebox{\linewidth}{!} {
+\begin{tabular}{lrrrrrl}
+  \hline
+& & Estimate & Error & $t$-value & \multicolumn{2}{l}{approx $p$-value} \\   \hline
+\\[-8pt]
+\bf design& Intercept & -3.11 & 0.41 & -7.61 & 0.00 & ***\\ 
+&  density & 0.05 & 0.57 & 0.09 & 0.93 \\ 
+&  dotplot & 0.46 & 0.57 & 0.81 & 0.42\\ 
+&  histogram & 0.06 & 0.58 & 0.11 & 0.91 \\ [2pt]
+\multicolumn{2}{l}{\bf covariates}\\
+&  d & 1.77 & 0.34 & 5.18 & 0.00 & ***\\ 
+&  n1$_{100}$ & 1.84 & 0.19 & 9.73 & 0.00 & ***\\ 
+&  ratio & 0.84 & 0.33 & 2.56 & 0.01 & **\\ 
+\multicolumn{2}{l}{\bf interactions}\\
+&  density:d & -0.60 & 0.47 & -1.27 & 0.20 \\ 
+&  dotplot:d & 0.03 & 0.46 & 0.07 & 0.95  \\ 
+&  histogram:d & -0.28 & 0.48 & -0.59 & 0.56 \\ [1pt]
+&  density:n1$_{100}$ & -0.10 & 0.26 & -0.36 & 0.72\\ 
+&  dotplot:n1$_{100}$ & -0.77 & 0.26 & -3.00 & 0.00 & ***\\ 
+&  histogram:n1$_{100}$ & -0.69 & 0.26 & -2.64 & 0.01 & ** \\ [1pt]
+&  density:ratio & -0.28 & 0.47 & -0.59 & 0.56 \\ 
+&  dotplot:ratio & -1.01 & 0.46 & -2.19 & 0.03 & * \\ 
+&  histogram:ratio & -0.28 & 0.47 & -0.59 & 0.55 \\ 
+   \hline
+\\[-5pt]
+   \multicolumn{5}{l}{Signif. codes:  0 `***' 0.001 `**' 0.01 `*' 0.05 `.' 0.1}
+\end{tabular}}
+\end{center}
+\caption{\label{tbl:power2}Results from a Generalized Linear Mixed Effects Model of power of a  lineup given design and all two-way interactions with effect size $d$, group size $n_1$, and ratio to second to first group size.
+Note that for all terms involving $n_1$ estimate and error are multiplied by a factor of 100. 
+Reference group for all effects are boxplot designs. The model fit is based on 2513 lineup evaluations by 208 participants. }
+\end{table}
 
@@
 \end{center}
-\caption{\label{tbl:speed2}Results from a Generalized Linear Mixed Effects Model of (log) time to answer a lineup given design and all two-way interactions with effect size $d$, group size $n_1$, and ratio to second to first group size.
+\caption{\label{tbl:speed2}Results from a Linear Mixed Effects Model of (log) time to answer a lineup given design and all two-way interactions with effect size $d$, group size $n_1$, and ratio to second to first group size.
 Estimate and error for all terms involving $n_1$ are multiplied by a factor of 100 for comparability.

### 2012-03-30 experiment.tex
table back in

@@
 
-Time taken to answer was log transformed before the modeling process to de-emphasize the impact of very large values (up to 500 sec).
+Time taken to answer was log transformed before the modeling process to de-emphasize the impact of very large values (up to 500 seconds).
 The findings are consistent with correctness - time taken shows big differences between polar and Euclidean charts.
-The average of the amount of time spent on a Euclidean lineup is 3.53 minutes while it was 4.07 minutes for a polar lineup.
+The average amount of time spent on a Euclidean lineup is $e^{3.53} = 34.1$ seconds compared to $e^{4.07}=58.5$ seconds for a polar lineup.
 The reference lines seem to increase the evaluation time, but not significantly.

### 2012-03-30 conclusions.tex
rearranged paper
some more on confidence levels in experiment 2
possible discussion points for conclusions

@@
-\section{Conclusions}
+\section{Conclusions}~\label{conclusions}
 
 Could say something about adding a visual pleasure rating to the results, rate how attractive the display is on a scale of 1-5.
+
+Familiarity with the design?
+
+Assessing confidence leads to interesting discrepancies between perceived versus actual accuracy.
+In both examples, the only significant increases in confidence were not indicative of an increase in actual performance, in one case even the opposite happened.
+
+Discussion of demographics: both studies have highly educated participants; turk5 slightly more men.
+there's a gender * design interaction - did not include in the model write-up, though.
+too tired for today
+turk4 has almost twice as many men.
+
+cross-tabulation of age vs education gives odd results - 108 participants between 18-25 have a graduate degree?

### 2012-03-30 experiment.tex
Bonferroni adjustment for fig 8

@@
 
+%Note that we excluded demographic information about individuals (i.e., age, gender and education level), as none of these covariates exhibited any significant relationship with accuracy while adding to model complexity.
+
 \begin{figure}[htbp] %  figure placement: here, top, bottom, or page
@@
    \caption{Power results for two competing designs: polar versus a Euclidean, with and without a reference line.
-Error bars indicate 95\% confidence intervals.
+Error bars indicate 95\% Bonferroni adjusted confidence intervals.
 Plots are facetted by percentage of data used.
@@
 Included are two-way interactions with sample size and shifts in wind direction (offset).
-Euclidean designs without a reference line at offset 0 are used as
+Euclidean designs without a reference line at offset 0 are used as baseline.
  Results are based on  976 lineup evaluations by 115 participants. }

### 2012-03-30 code/code-for-figures.R

@@ -129,3 +129,26 @@ ddply(conc.res, .(n1,n2,d,test_param), summarize,
 
 
 xtabs(~ pic_name + responsedata=conc.res)
+
+
+
+nf <- ldply(c(0,90,180,270), function(x) {
+	flights_sample$offset <- x
+	flights_sample$WindDirDegrees <- (flights_sample$WindDirDegrees+x)%%360
+	flights_sample
+})
+
+ggplot(nf, aes(x = (WindDirDegrees-1 + X)%%360, fill = factor(nf, levels = c(0:8)))) +
+geom_bar(binwidth = 10, position="fill", alpha=0.8) +
+scale_x_continuous(limits = c(0, 360), breaks=seq(0, 360, by=45), labels=c("N", "NE","E","SE","S","SW","W","NW","N")) + geom_hline(yintercept=0.5, size = .5, alpha=0.5, colour = "white") + opts(legend.position="none", plot.margin=unit(c(0,0,0,0), "cm")) +
+seq7rev + labs(x = " ", y = " ") +facet_grid(.~offset)
+ggsave("euclid-offset.pdf")
+
+ggplot(nf, aes(x = (WindDirDegrees-1 + X)%%360, fill = factor(nf, levels = c(0:8)))) +
+geom_bar(binwidth = 10, position="fill") +
+scale_x_continuous(limits = c(0, 360), breaks=seq(0, 360, by=45), labels=c("N", "NE","E","SE","S","SW","W","NW","N")) +
+coord_polar(start = 0) + geom_line(aes(x=xx, y=yy), size = .5, alpha=0.5, colour = "white", inherit.aes=FALSE, data=refdf) + opts(legend.position="none", plot.margin=unit(c(0,0,0,0), "cm")) +
+seq7rev + labs(x = " ", y = " ")  +facet_grid(.~offset)
+ggsave("polar-offset.pdf")
+
+

### 2012-03-30 experiment.tex
fixed consistency in references

@@
 The thin lines represent power for individuals.
-What can be seen is the different impact of the offset by design: while  an offset of 270 degrees (the 'mountain' pattern) has the highest power in Euclidean charts, it comes out worst in polar charts.
+What can be seen is the different impact of the offset by design: while  an offset of 270 degrees (the `mountain' pattern) has the highest power in Euclidean charts, it comes out worst in polar charts.
 

### 2012-03-31 experiment.tex
restructured a bit, filled in missing pieces

@@
 \section{Results and Discussion}~\label{results}
+%
+\subsection{Wind Direction and Airport Efficiency}
+Figures \ref{fig:time} and \ref{fig:conf} give an overview of the relationship between the three response variables: accuracy, speed, and confidence level.
 
-\subsection{Wind Direction and Airport Efficiency}
-Figure \ref{fig:time} shows time taken until a decision was made in histograms, facetted by design and colored by correctness of answer.
+Figure \ref{fig:time} shows histogram of the time participants needed to make a decision on each lineup.
+Correctness of answers is shown by color.
 Because of the skewness of the distributions, times were log-transformed.
-Polar charts take on average more time to answer, but are answered with much lower precision.
-Figure \ref{fig:conf} shows two barcharts of confidence levels by task, again coloring is used for correctness of answers.
-Euclidean displays lead to a very bimodal distribution of confidence: participants are either very sure or not sure at all of their answer.
-Confidence levels in polar charts are distributed much more uniformly.
-
-The perceptual tasks involved in decoding the designs fit into the general framework of the piechart versus barchart discussion.
-However, we are not dealing with assessing angles, which is known to be a perceptual harder task than assessments of heights or widths in bars, see e.g.
-\cite{cleveland:1984, robbins:2004, Kosslyn:2006, few:2009}.
-Instead, the perceptual tasks consists of comparisons along a common axis (in the barcharts) and a common origin (in the polar charts).
-The difference in designs is therefore based on how well we are able to judge deviations from a horizontal line compared to deviations from a circle.
-Based on this, the charts with added reference lines provide us with exactly the frame we compare with and should, therefore, be the `better' designs - either in speed or accuracy.
+Polar charts take on average more time to answer, but are answered with much lower accuracy.
+The average amount of time spent on a Euclidean lineup is $e^{3.53} = 34.1$ seconds compared to $e^{4.07}=58.5$ seconds for a polar lineup.
+This is in stark contrast to accuracy: 76.92\% of the Euclidean lineups shown resulted in the a correct identification of the real data while only 20.31\% of the polar charts were correct.
 
@@
 
+Figure \ref{fig:conf} shows two barcharts of confidence levels by task, again coloring is used for correctness of answers.
+Euclidean displays lead to a very bimodal distribution of confidence: participants are either very sure or not sure at all of their answer.
+Confidence levels in polar charts are distributed much more uniformly.
 \begin{figure}[hbtp] %  figure placement: here, top, bottom, or page
@@
 \end{figure}
+
+The perceptual tasks involved in decoding the designs fit into the general framework of the piechart versus barchart discussion.
+However, we are not dealing with assessing angles, which is known to be a perceptual harder task than assessments of heights or widths in bars, see e.g.
+\cite{cleveland:1984, robbins:2004, Kosslyn:2006, few:2009}.
+Instead, the perceptual tasks consist of comparisons along a common axis (in the barcharts) and a common origin (in the polar charts).
+The difference in designs is therefore based on how well we are able to judge deviations from a horizontal line compared to deviations from a circle.
+Based on this, the charts with added reference lines provide us with exactly the frame we compare with and should, therefore, be the `better' designs - either in speed or accuracy.
 
@@
 
-For evaluating the results, we are using a generalized linear mixed effects modeling  approach \cite{pinheiro:2000} for all three of our response values, using the {\t R} package {\tt lme4} \cite{bates:2011}.
+For evaluating the results, we are using a generalized linear mixed effects modeling  approach \cite{pinheiro:2000} for each one of the three response values, using the {\t R} package {\tt lme4} \cite{bates:2011}.
 For power predictions (cf.
@@
 To adjust for individuals' ability, a random intercept was included in the model.
-
-%Note that we excluded demographic information about individuals (i.e., age, gender and education level), as none of these covariates exhibited any significant relationship with accuracy while adding to model complexity.
-
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-
-\begin{tabular}{cl}
-\phantom{\includegraphics[width=0.03\linewidth]{Polar_Line.pdf}} & \vspace{-0.035in} \multirow{10}{*}{\hspace{-0.25in}\includegraphics[width=0.9\linewidth]{turk4-designs.pdf}} \\
-\includegraphics[width=0.05\linewidth]{Polar_Line.pdf} \\
-\includegraphics[width=0.05\linewidth]{Polar_NoLine.pdf} \\
-\includegraphics[width=0.043\linewidth]{Euclidian_Line.pdf} \\
-\includegraphics[width=0.043\linewidth]{Euclidian_NoLine.pdf}\\
-\phantom{\includegraphics[width=0.015\linewidth]{Polar_Line.pdf}}\\
-\includegraphics[width=0.05\linewidth]{Polar_Line.pdf} \\
-\includegraphics[width=0.05\linewidth]{Polar_NoLine.pdf} \\
-\includegraphics[width=0.043\linewidth]{Euclidian_Line.pdf} \\
-\includegraphics[width=0.043\linewidth]{Euclidian_NoLine.pdf}\\
-  \end{tabular} 
-  \vspace{0.1in}
-   \caption{Power results for two competing designs: polar versus a Euclidean, with and without a reference line.
-Error bars indicate 95\% Bonferroni adjusted confidence intervals.
-Plots are facetted by percentage of data used.
-}
-   \label{fig:treatment}
-\end{figure}
-
-Overall, the results show huge differences in the power of designs between polar charts and Euclidean: Euclidean designs are significantly more powerful than polar charts, particularly so with small sample sizes.
-76.92\% of the Euclidean lineups shown resulted in the a correct identification of the real data while only 20.31\% of the polar charts were correct.
- The reference line has surprisingly little influence, but it helps more for  polar charts than for Euclidean charts,
-An increase in sample size has positive impact on power (see figure \ref{fig:treatment} for an overview).
-Polar charts need a much bigger sample size to see an increase in power -- only at about 24\% of the original data do we see about the same power as for Euclidean charts of a sample size of 2\%.
-The  changes in offset are significant -- interestingly, borderline behavior (90 and 270 degrees) does not show a difference between polar and Euclidean charts, whereas an inversion of the wave pattern (first up, then down), does show a difference.
-The power of Euclidean lineups suffers significantly  from this inversion whereas power of polar charts is unaffected.
-This is an unexpected finding, but is consistent throughout different lineups in the data.
-Figure \ref{fig:power} summarizes the results from the model: power predictions ($y$ axis) are shown by sample size ($x$ axis).
-The thick lines show average power by design for different shifts in wind direction.
-The thin lines represent power for individuals.
-What can be seen is the different impact of the offset by design: while  an offset of 270 degrees (the `mountain' pattern) has the highest power in Euclidean charts, it comes out worst in polar charts.
-
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-   \includegraphics[width=\linewidth]{predict-power} 
-   \caption{Predicted Power of designs.
-The thin lines and the points on $y$ axis show  variability due to individuals' abilities.
-The saturated lines show average predicted power for each of the designs.}
-   \label{fig:power}
-\end{figure}
-
-Time taken to answer was log transformed before the modeling process to de-emphasize the impact of very large values (up to 500 seconds).
-The findings are consistent with correctness - time taken shows big differences between polar and Euclidean charts.
-The average amount of time spent on a Euclidean lineup is $e^{3.53} = 34.1$ seconds compared to $e^{4.07}=58.5$ seconds for a polar lineup.
-The reference lines seem to increase the evaluation time, but not significantly.
-An increase in sample size decreases evaluation time, The inversion of the wave pattern at 180 degrees leads to a significant increase in time for Euclidean charts, but not for polar charts (at least not significantly).
-
- Confidence levels are measured on a five point scale --- they are subjective assessments by the participant `how certain are you'.
-We model confidence using the same model structure as before, i.e.
-using sample size and offset as covariates and including up to two-way interactions,
-While there is a difference between the designs - participants reported higher confidence in dealing with the Euclidean lineups -- this is only  a trend (i.e.
-not significant at 5\% but below 10\%).
-The only significant effect on reported confidence level is the use of reference lines in polar coordinates: participants report an increase in confidence (0.31 $\pm$ 0.13, $p$ value = 0.01) when using the reference line.
-This, however, does not translate to a significant increase in accuracy of results, as we have seen before.
-
- Euclidean coordinates resulted in a significantly more accurate identification of the real data set in significantly shorter time.
 
@@
 
+%Note that we excluded demographic information about individuals (i.e., age, gender and education level), as none of these covariates exhibited any significant relationship with accuracy while adding to model complexity.
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+
+\begin{tabular}{cl}
+\phantom{\includegraphics[width=0.03\linewidth]{Polar_Line.pdf}} & \vspace{-0.035in} \multirow{10}{*}{\hspace{-0.25in}\includegraphics[width=0.9\linewidth]{turk4-designs.pdf}} \\
+\includegraphics[width=0.05\linewidth]{Polar_Line.pdf} \\
+\includegraphics[width=0.05\linewidth]{Polar_NoLine.pdf} \\
+\includegraphics[width=0.043\linewidth]{Euclidian_Line.pdf} \\
+\includegraphics[width=0.043\linewidth]{Euclidian_NoLine.pdf}\\
+\phantom{\includegraphics[width=0.015\linewidth]{Polar_Line.pdf}}\\
+\includegraphics[width=0.05\linewidth]{Polar_Line.pdf} \\
+\includegraphics[width=0.05\linewidth]{Polar_NoLine.pdf} \\
+\includegraphics[width=0.043\linewidth]{Euclidian_Line.pdf} \\
+\includegraphics[width=0.043\linewidth]{Euclidian_NoLine.pdf}\\
+  \end{tabular} 
+  \vspace{0.1in}
+   \caption{Power results for two competing designs: polar versus a Euclidean, with and without a reference line.
+Error bars indicate 95\% Bonferroni adjusted confidence intervals.
+Plots are facetted by percentage of data used.
+}
+   \label{fig:treatment}
+\end{figure}
+
+Overall, the results show huge differences in the power of designs between polar charts and Euclidean: Euclidean designs are significantly more powerful than polar charts, particularly so with small sample sizes.
+ The reference line has surprisingly little influence, but it helps more for  polar charts than for Euclidean charts,
+An increase in sample size has a positive impact on power (see figure \ref{fig:treatment} for an overview).
+Polar charts need a much bigger sample size to see an increase in power -- only at about 24\% of the original data do we see about the same power as for Euclidean charts of a sample size of 2\%.
+The  changes in offset are significant -- interestingly, borderline behavior (90 and 270 degrees) does not show a difference between polar and Euclidean charts, whereas an inversion of the wave pattern (first up, then down), does show a difference.
+The power of Euclidean lineups suffers significantly  from this inversion whereas power of polar charts is unaffected.
+This is an unexpected finding, but is consistent throughout different lineups in the data.
+Figure \ref{fig:power} summarizes the results from the model: power predictions ($y$ axis) are shown by sample size ($x$ axis).
+The thick lines show average power by design for different shifts in wind direction.
+The thin lines represent power for individuals.
+What can be seen is the different impact of the offset by design: while  an offset of 270 degrees (the `mountain' pattern) has the highest power in Euclidean charts, it comes out worst in polar charts.
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+   \includegraphics[width=\linewidth]{predict-power} 
+   \caption{Predicted Power of designs.
+The thin lines and the points on $y$ axis show  variability due to individuals' abilities.
+The saturated lines show average predicted power for each of the designs.}
+   \label{fig:power}
+\end{figure}
+
+\begin{figure*}[hbtp] %  figure placement: here, top, bottom, or page
+   \centering
+   \includegraphics[width=0.9\linewidth]{power-exp2} 
+   \caption{Overview of power predictions for the four different designs.
+The fully saturated thick lines show average predicted power for each of the designs facetted by size of the red group (top to bottom) and relative size of the blue group to the red group (left to right).
+Thin lines represent variability due to subject-specific abilities.
+}
+   \label{fig:power2}
+\end{figure*}
+
+Time taken to answer was log transformed before the modeling process to de-emphasize the impact of very large values (up to 500 seconds).
+The findings are consistent with correctness - time taken shows big differences between polar and Euclidean charts.
+The reference lines seem to increase the evaluation time, but not significantly.
+An increase in sample size decreases evaluation time, The inversion of the wave pattern at 180 degrees leads to a significant increase in time for Euclidean charts, but not for polar charts (at least not significantly).
+Table \ref{tbl:time} shows an overview of the model parameters and estimates.
+
 \begin{table}[ht]
@@
 \end{table}
+
+ Confidence levels are measured on a five point scale --- they are subjective assessments by the participant `how certain are you'.
+We model confidence using the same model structure as before, i.e.
+using sample size and offset as covariates and including up to two-way interactions,
+While there is a difference between the designs - participants reported higher confidence in dealing with the Euclidean lineups -- this is only  a trend (i.e.
+not significant at 5\% but below 10\%).
+The only significant effect on reported confidence level is the use of reference lines in polar coordinates: participants report an increase in confidence (0.31 $\pm$ 0.13, $p$ value = 0.01) when using the reference line.
+This, however, does not translate to a significant increase in accuracy of results, as we have seen before.
+
+ Euclidean coordinates resulted in a significantly more accurate identification of the real data set in significantly shorter time.
 

### 2012-03-31 experiment-two.tex
small changes
Merge branch 'master' of github.com:heike/comparing-designs
Conflicts:
	setup.tex

@@
 % !TEX root = comparison.tex
-
-\subsection{Comparing Distributions}
-\begin{figure*}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-   \includegraphics[width=\linewidth]{power-exp2} 
-   \caption{Overview of power predictions for the four different designs.
-The fully saturated thick lines show average predicted power for each of the designs facetted by size of the red group (top to bottom) and relative size of the blue group to the red group (left to right).
-Thin lines represent variability due to subject-specific abilities.
-}
-   \label{fig:power2}
-\end{figure*}
+\subsection{Comparing Distribution Centers}
 
@@
 The two highly summarized designs benefit the most from an increase in sample size: both density charts and boxplots profit significantly more than histograms and dotplots.
-This might, again, be a hint towards a trade-off between visual complexity and amount of information shown as basis for decisions.
+This might, again, be a hint towards a trade-off between visual complexity and amount of information shown as the basis for making decisions.
 
@@
 This means that we need to interpret those  estimates as average change if the group size is increased by 100 elements.
-The results of this model are similar to the previous: evaluation speeds needed for the four different designs is very close, but  histograms take on average a significantly longer time ($e^{0.23} = 1.26$ seconds) to evaluate than lineups of boxplots.
-Where an increase of sample size led to an increase of accuracy before, we now see that it increases time for evaluation of boxplot slightly, but not significantly so.
-What is interesting is that evaluation time decreases significantly with an increase in sample size, particularly for dotplots, but also for density charts.
-This is counter-intuitive to our previous argument on dotplots being affected by visual complexity the most.
-It seems, that observers of dotplot lineups come to a conclusion faster as sample size increases, albeit more often to  the wrong conclusion.
-Rather than visual complexity, this might be indicative that over-plotting is problematic in dotplots.
-While we tried to accommodate for over-plotting by making use of alpha-blending and jittering, this might not have been sufficient for the large sample size.
 
@@
 
-Our third response variable is again the subjective assessment of confidence level.
-On average participants reported a confidence in their results of  3.41 (on a five point scale with 1=lowest and 5=highest), with a standard deviation of 0.149.
-Modeling confidence levels with a linear mixed effects model with the same structure as before resulted in only one significant effect beyond the intercept: participants report significantly higher confidence for increased number of  points in dotplot lineups.
-On average an increase of 100 points leads to an increase in confidence of 0.22 (with a standard error of 0.082).
-This finding is significant with a $p$-value of 0.007.
-Together with the previous results, we can tell that something worth of further investigation happens in the perception of dotplots as sample sizes increase.
+The results of this model are similar to the previous: evaluation speeds needed for the four different designs is very close, but  histograms take on average a significantly longer time ($e^{0.23} = 1.26$ seconds) to evaluate than lineups of boxplots.
+Where an increase of sample size led to an increase of accuracy before, we now see that it increases time for evaluation of boxplot slightly, but not significantly so.
+What is interesting is that evaluation time decreases significantly with an increase in sample size, particularly for dotplots, but also for density charts.
+This is counter-intuitive to our previous argument on dotplots being affected by visual complexity the most.
+It seems, that observers of dotplot lineups come to a conclusion faster as sample size increases, albeit more often to  the wrong conclusion.
+Rather than visual complexity, this might be indicative that over-plotting is problematic in dotplots.
+While we tried to accommodate for over-plotting by making use of alpha-blending and jittering, this might not have been sufficient for the large sample size.
 
@@
 
+Our third response variable is again the subjective assessment of confidence level.
+On average participants reported a confidence in their results of  3.41 (on a five point scale with 1=lowest and 5=highest), with a standard deviation of 0.149.
+Modeling confidence levels with a linear mixed effects model with the same structure as before resulted in only one significant effect beyond the intercept: participants report significantly higher confidence for increased number of  points in dotplot lineups.
+On average an increase of 100 points leads to an increase in confidence of 0.22 (with a standard error of 0.082).
+This finding is significant with a $p$-value of 0.007.
+Together with the previous results, we can tell that something worth of further investigation happens in the perception of dotplots as sample sizes increase.
+
 Figure \ref{fig:conf-margins} shows barcharts of confidence assessments by design.

### 2012-03-31 conclusions.tex
conclusions

@@
+% !TEX root = comparison.tex
 \section{Conclusions}~\label{conclusions}
+Lineup charts allow us to assess and quantify the strength of a visual finding.
+In this paper we introduce the concept of  the power of a lineup, and show its use as a measure to evaluate competing designs.
+Power of a lineup can be modeled in a logistic regression model with subject-specific random effects, if observers were exposed and provided feedback to multiple lineups.
+
+We describe two MTurk studies evaluating different designs.
 

### 2012-03-31 conclusions.tex
conclusions, mainly

--- 2012-03-29 conclusions.tex
@@
-\section{Conclusions}
+% !TEX root = comparison.tex
+\section{Conclusions}~\label{conclusions}
+Lineup charts allow us to assess and quantify the strength of a visual finding.
+In this paper we introduce the concept of  the power of a lineup, and show its use as a measure to evaluate competing designs.
+Power of a lineup can be modeled in a logistic regression model with subject-specific random effects, if observers were exposed and provided feedback to multiple lineups.
 
-Could say something about adding a visual pleasure rating to the results, rate how attractive the display is on a scale of 1-5.
+We describe two MTurk studies, where we use lineups to  evaluate competing designs.
+By varying the experimental conditions we are able to compare designs under slightly different  circumstances, which allows us to come to more general conclusions about weaknesses and strengths of one design over other designs.
+While the MTurk results supported our choice of dotplots in the very small study that triggered our investigation, the results also brought the message of visual complexity home: showing much more information than necessary for a particular decision is detrimental to accuracy rather than supportive.
+
+Assessing confidence leads to interesting discrepancies between perceived versus actual accuracy.
+In both examples, the only significant increases in confidence were not indicative of an increase in actual performance, in one case even the opposite occurred.
+Besides accuracy and speed, there is a variety of other aspects of the design, which could be collected in  the same framework of  MTurk studies, such as ease of use, familiarity with a particular type of design, or a visual pleasure rating of the lineups  (on a scale of one to five).

### 2012-03-31 experiment.tex
conclusions, mainly
included the pairwise comparison

@@
 \subsection{Wind Direction and Airport Efficiency}
+
 Figures \ref{fig:time} and \ref{fig:conf} give an overview of the relationship between the three response variables: accuracy, speed, and confidence level.
@@
 
+Figure \ref{fig:treatment} shows a comparison of power for the four competing designs.
+95\% confidence intervals are  Bonferroni adjusted for multiple testing.
+The bottom two confidence intervals in each panel show a 95\% confidence interval of a direct comparison of the Euclidean design versus the polar design with and without a reference line.
+With the exception of the 2\% sample none of those confidence intervals include the zero, indicating a significantly higher power for the Euclidean design than for the polar design.
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+
+\begin{tabular}{cl}
+\phantom{\includegraphics[width=0.03\linewidth]{Polar_Line.pdf}} & \vspace{-0.035in} \multirow{10}{*}{\hspace{-0.5in}\includegraphics[width=0.9\linewidth]{turk4-designs.pdf}} \\
+\includegraphics[width=0.045\linewidth]{Polar_Line.pdf} \\
+\includegraphics[width=0.045\linewidth]{Polar_NoLine.pdf} \\
+\includegraphics[width=0.043\linewidth]{Euclidian_Line.pdf} \\
+\includegraphics[width=0.043\linewidth]{Euclidian_NoLine.pdf}\\
+\phantom{\includegraphics[width=0.13\linewidth]{Polar_Line.pdf}}\\
+\includegraphics[width=0.045\linewidth]{Polar_Line.pdf} \\
+\includegraphics[width=0.045\linewidth]{Polar_NoLine.pdf} \\
+\includegraphics[width=0.043\linewidth]{Euclidian_Line.pdf} \\
+\includegraphics[width=0.043\linewidth]{Euclidian_NoLine.pdf}\\
+\phantom{\includegraphics[width=0.13\linewidth]{Polar_Line.pdf}}\\
+  \end{tabular} 
+  \vspace{0.1in}
+   \caption{Power results for two competing designs: polar versus a Euclidean, with and without a reference line; facetting indicates sample size (as percentage of original data).
+Error bars indicate single Bonferroni adjusted 95\% confidence intervals.
+The two confidence intervals at the bottom of each panel shows a power comparison of Euclidean versus polar design according to eqn (\ref{eq:comp}).
+}
+   \label{fig:treatment}
+\end{figure}
+
 %Most generally this experiment can be paralleled to the pie chart versus barchart discussion \cite{***}. 
@@
 
-For evaluating the results, we are using a generalized linear mixed effects modeling  approach \cite{pinheiro:2000} for each one of the three response values, using the {\t R} package {\tt lme4} \cite{bates:2011}.
+However, all of these considerations are based on the - rather strong - assumption that all individuals have the same ability to detect the data plot from a lineup.
+In order to allow for individual differences in visual ability,
+ we use a generalized linear mixed effects modeling  approach \cite{pinheiro:2000} for each of the three response values, using the {\tt R} package {\tt lme4} \cite{bates:2011}. 
 For power predictions (cf.
@@
 
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-
-\begin{tabular}{cl}
-\phantom{\includegraphics[width=0.03\linewidth]{Polar_Line.pdf}} & \vspace{-0.035in} \multirow{10}{*}{\hspace{-0.25in}\includegraphics[width=0.9\linewidth]{turk4-designs.pdf}} \\
-\includegraphics[width=0.05\linewidth]{Polar_Line.pdf} \\
-\includegraphics[width=0.05\linewidth]{Polar_NoLine.pdf} \\
-\includegraphics[width=0.043\linewidth]{Euclidian_Line.pdf} \\
-\includegraphics[width=0.043\linewidth]{Euclidian_NoLine.pdf}\\
-\phantom{\includegraphics[width=0.015\linewidth]{Polar_Line.pdf}}\\
-\includegraphics[width=0.05\linewidth]{Polar_Line.pdf} \\
-\includegraphics[width=0.05\linewidth]{Polar_NoLine.pdf} \\
-\includegraphics[width=0.043\linewidth]{Euclidian_Line.pdf} \\
-\includegraphics[width=0.043\linewidth]{Euclidian_NoLine.pdf}\\
-  \end{tabular} 
-  \vspace{0.1in}
-   \caption{Power results for two competing designs: polar versus a Euclidean, with and without a reference line.
-Error bars indicate 95\% Bonferroni adjusted confidence intervals.
-Plots are facetted by percentage of data used.
-}
-   \label{fig:treatment}
-\end{figure}
-
 Overall, the results show huge differences in the power of designs between polar charts and Euclidean: Euclidean designs are significantly more powerful than polar charts, particularly so with small sample sizes.
  The reference line has surprisingly little influence, but it helps more for  polar charts than for Euclidean charts,
-An increase in sample size has a positive impact on power (see figure \ref{fig:treatment} for an overview).
+An increase in sample size has a positive impact on power.
 Polar charts need a much bigger sample size to see an increase in power -- only at about 24\% of the original data do we see about the same power as for Euclidean charts of a sample size of 2\%.

### 2012-03-31 conclusions.tex
tweaked conclusions

@@
 \section{Conclusions}~\label{conclusions}
-Lineup charts allow us to assess and quantify the strength of a visual finding.
+
+In this paper we have reported on results of using lineups for evaluating plot designs using a measure called power.
+A logistic regression model with subject-specific random effects is used to obtain a model for estimating power, and account for individual subject abilities.
+
+%Lineup charts allow us to assess and quantify the strength of a visual finding.
 In this paper we introduce the concept of  the power of a lineup, and show its use as a measure to evaluate competing designs.
@@
 By varying the experimental conditions we are able to compare designs under slightly different circumstances, which allows us to come to more general conclusions about weaknesses and strengths of one design over other designs.
-While the MTurk results supported our choice of dotplots in the very small study that triggered our investigation, the results also brought the message of visual complexity home: showing much more information than necessary for a particular decision is detrimental to accuracy rather than supportive.
+Euclidean charts are significantly more powerful than polar charts.
+The MTurk results also supported our choice of dotplots in the very small samples that triggered our investigation, but the results elucidated the visual complexity dilemma: showing much more information than necessary for a particular decision is detrimental to accuracy rather than supportive.
 
-Assessing confidence leads to interesting discrepancies between perceived versus actual accuracy.
+Assessing the self-reported confidence leads to interesting discrepancies between perceived versus actual accuracy.
 In both examples, the only significant increases in confidence were not indicative of an increase in actual performance, in one case even the opposite occurred.
-Besides accuracy and speed, there is a variety of other aspects of the design, which could be collected in  the same framework of  MTurk studies, such as ease of use, familiarity with a particular type of design, or a visual pleasure rating of the lineups  (on a scale of one to five).
+Besides accuracy and speed, there is a variety of other aspects of the design, which could be collected in  the same framework of  MTurk studies, such as ease of use, familiarity with a particular type of design, or a visual pleasure rating of the lineups  (on a scale of 1-5).

### 2012-03-31 experiment.tex
tweaked results

@@
    \includegraphics[width=\linewidth]{conf-answer} 
-   \caption{Barcharts of (subjective) confidence in answering correctly for each lineup.
+   \caption{Barcharts of self-reported confidence in answering correctly for each lineup.
 A strongly bimodal distribution is apparent, while confidence levels for polar charts are more uniform.
@@
 Error bars indicate single Bonferroni adjusted 95\% confidence intervals.
-The two confidence intervals at the bottom of each panel shows a power comparison of Euclidean versus polar design according to eqn (\ref{eq:comp}).
-}
+The two confidence intervals at the bottom of each panel shows a pairwise power comparison of Euclidean versus polar design according to eqn (\ref{eq:comp}).
+Differences from zero (vertical grey line) are of interest.}
    \label{fig:treatment}

### 2012-03-31 experiment-two.tex
tweaked results
typos

@@
    \includegraphics[width=\linewidth]{conf-exp2} 
-   \caption{Barcharts of marginal distributions of subjective confidence assessment by design. The distributions display prominent differences.}
+   \caption{Barcharts of marginal distributions of self-reported confidence by design. The distributions display prominent differences.}
    \label{fig:conf-margins}
@@
 This finding is significant with a $p$-value of 0.007.
-Together with the previous results, we can tell that something worth of further investigation happens in the perception of dotplots as sample sizes increase.
+Together with the previous results, we can tell that something worthy of further investigation happens in the perception of dotplots as sample sizes increase.
 
@@
 It is obvious that the distribution of confidence is very different from one design to the next.
-Boxplots and Dotplots show a similar pattern with an emphasis on high confidence levels, whereas histograms and density lineups have modes at a confidence level of three.
+Boxplots and dotplots show a similar pattern with an emphasis on high confidence levels, whereas histograms and density lineups have modes at a confidence level of three.
 One reason behind these differences in confidence distributions might be familiarity with the design - neither density charts nor histograms are particularly frequent displays for a general audience.

### 2012-03-31 experiment.tex
typos

@@
 Because of the skewness of the distributions, times were log-transformed.
-Polar charts take on average more time to answer, but are answered with much lower accuracy.
+Polar charts take on average more time to answer, and are answered with much lower accuracy.
 The average amount of time spent on a Euclidean lineup is $e^{3.53} = 34.1$ seconds compared to $e^{4.07}=58.5$ seconds for a polar lineup.
@@
 A strongly bimodal distribution is apparent, while confidence levels for polar charts are more uniform.
-There are only slight (and non significant) difference between confidence levels and correctness of results.}
+Confidence levels are not indicative of accuracy: none of the differences between confidence levels show a significant difference in accuracy.}
    \label{fig:conf}
@@
 The perceptual tasks involved in decoding the designs fit into the general framework of the piechart versus barchart discussion.
-However, we are not dealing with assessing angles, which is known to be a perceptual harder task than assessments of heights or widths in bars, see e.g.
+However, we are not dealing with assessing angles, which is known to be perceptually harder than assessments of heights or widths in bars, see e.g.
 \cite{cleveland:1984, robbins:2004, Kosslyn:2006, few:2009}.
@@
   \vspace{0.1in}
-   \caption{Power results for two competing designs: polar versus a Euclidean, with and without a reference line; facetting indicates sample size (as percentage of original data).
+   \caption{Power results for two competing designs: polar versus  Euclidean, with and without a reference line; facetting indicates sample size (as percentage of original data).
 Error bars indicate single Bonferroni adjusted 95\% confidence intervals.

### 2012-06-27 review/response.txt
revision

I would like to thank the reviewers for the great discussion of the paper, and a lot of helpful suggestions!

We have made the following changes:

All of the main suggestions are implemented:
 - in the presentation of figure 8, differences between power results are now presented using a Benjamini-Hochberg adjustment for the false discovery rate.
   Since BH adjustments are based on individual p-values (whereas the Bonferroni adjustment results in a global scaling of the intervals), it was necessary to change the presentation in the figure slightly - we now show the significant differences using a letter-based approach as suggested by H.-P. Piepho (2004).

 - the paper is re-organized and the two studies are now presented one after the other.

 - the discussion regarding pie vs barcharts is removed - we'd like to thank the reviewers for pointing out the additional references.

Changes based on suggestions of individual reviewers:
 - the blue scale for the pictures of the wind direction was darkened a bit, which makes the small levels better distinguishable both in print and pdf.

 - reporting the log: we indeed used log base 10 for figure 6 - we changed this picture to natural log for consistency.

 - Euclidean vs Cartesian coordinates - agreed, and changed

 - language concerning null hypothesis and alternative in 3.2 is fixed.

 - The scale for the self-reported confidence is included explicitly now.

 -  the design choices for the second study were based on designs commonly used; we have addressed this in the paper.

 - conclusion section is strengthened and also focuses on the results from the experiment rather than just their setup.

Discussion:

Q: Would it be more understandable and specific to characterize the power
  model as a hierarchical linear model with random intercept instead of a
  generalized linear model? If I understand correctly, the HLM with random
  intercept is specifically what they are using. If not, the authors can
  explain to the reviewers why I am wrong and not make the change.

A: What we are using in estimating power is actually a generalized linear model with a random intercept - since power has to be a value between 0 and 1, we are using a logistic regression instead of a linear model.

Q: The first study does not state clearly what constitutes "best" time for
  minutes between wheel events. You claim that it is clear that Seattle
  functions at its best with winds coming from the East but I'm not sure if
  that means that you just want the lowest number of minutes or if there is
  an optimal number. The hypotheses just address "differences" but not any
  specific time interval in general.

A: We were trying to avoid the phrase 'best' - as we are, indeed, just looking at efficiency ('its best' has been changed in the paper to 'most efficiently') - i.e. less time between wheel events is considered 'better'. (Note: Seattle has several runways, so a time of 0 between wheel events is not as unsafe as it might sound).

Hans-Peter Piepho (2004), An Algorithm for a Letter-Based Representation of All-Pairwise Comparisons, Journal of Computational and Graphical Statistics, 13(2), 456–466.

### 2012-06-27 review/comparison-review.tex
revision

--- 2012-03-29 comparison.tex
@@
 \usepackage[table]{xcolor}
+
+\renewcommand\floatpagefraction{1}
+
 %% We encourage the use of mathptmx for consistent usage of times font
@@
 %% indicate IEEE Member or Student Member in form indicated below
-\author{Lendie Follett, Heike Hofmann, Mahbubul Majumder, Dianne Cook}
+\author{Heike Hofmann, Lendie Follett, Mahbubul Majumder, Dianne Cook}
 \authorfooter{
@@
 \item
+ Heike Hofmann is an associate professor in Statistics at Iowa State University, E-mail: hofmann@iastate.edu.
+\item
  Lendie Follett is an undergraduate student in Statistics at Iowa State University, E-mail: lfollett@iastate.edu.
-\item
- Heike Hofmann is an associate professor in Statistics at Iowa State University, E-mail: hofmann@iastate.edu.
 \item
@@
 design. The first study compares the choice of coordinate system:
-polar versus Euclidean coordinates. The results show strong support in
-favor of Euclidean coordinates in finding fast and accurate answers to
+polar versus cartesian coordinates. The results show strong support in
+favor of cartesian coordinates in finding fast and accurate answers to
 spotting patterns. The second study is aimed at finding shift
@@
 simulated data to evaluate the plot designs under controlled
-conditions.  Amazon Turk is used to conduct the studies.  The lineups
+conditions.  Amazon Mechanical Turk (MTurk) is used to conduct the studies.  The lineups
 provide an effective mechanism for objectively evaluating plot

### 2012-06-27 review/conclusions.tex
revision

% !TEX root = comparison-review.tex
\section{Conclusions}~\label{conclusions}

%In this paper we have reported on results of using lineups for evaluating plot designs using a measure called power.  A logistic regression model with subject-specific random effects is used to obtain a model for estimating power, and account for individual subject abilities. 
%
%%Lineup charts allow us to assess and quantify the strength of a visual finding. In this paper we introduce the concept of  the power of a lineup, and show its use as a measure to evaluate competing designs. Power of a lineup can be modeled in a logistic regression model with subject-specific random effects, if observers were exposed and provided feedback to multiple lineups. 
%
%We describe two MTurk studies, where we use lineups to  evaluate competing designs. By varying the experimental conditions we are able to compare designs under slightly different circumstances, which allows us to come to more general conclusions about weaknesses and strengths of one design over other designs.  Cartesian charts are significantly more powerful than polar charts. The MTurk results also supported our choice of dotplots in the very small samples that triggered our investigation, but the results elucidated the visual complexity dilemma: showing much more information than necessary for a particular decision is detrimental to accuracy rather than supportive.
%
%Assessing the self-reported confidence leads to interesting discrepancies between perceived versus actual accuracy. In both examples, the only significant increases in confidence were not indicative of an increase in actual performance, in one case even the opposite occurred. 
%
%Besides accuracy and speed, there is a variety of other aspects of the design, which could be collected in  the same framework of  MTurk studies, such as ease of use, familiarity with a particular type of design, or a quantitative visual pleasure rating of the lineups.

In this paper we are investigating lineups for evaluating competing plot designs based on their power modeled by a logistic regression with subject-specific random intercepts. 
Lineups provide a powerful  tool for evaluating different designs in the framework of the data. We can show in two MTurk studies, that cartesian charts are significantly more powerful than polar charts. The  results also supported our choice of dotplots in the very small samples that triggered our investigation, but  elucidated the visual complexity dilemma: showing much more information than necessary for a particular decision is detrimental to accuracy rather than supportive.

While we --theoretically-- can't show the same data plot to a single individual, it does make in practice a quite convincing tool to do so, e.g. in the informal setting of a classroom or in interactions with collaborators, when we do not intend to collect data for further investigation of a problem. 
Using the MTurk service gives us fast feedback -- both of the studies were completed within hours of their availability -- but comes at the price that we have to deal with the occasional `gamer'. For future studies we are planning on having a required entrance  test to try to avoid those. Besides accuracy and speed, there is a variety of other aspects of the design, which could be collected in  the same framework of  MTurk studies, such as ease of use, familiarity with a particular type of design, or a quantitative visual pleasure rating of the lineups.

Different visual abilities of individuals make up a significant amount of variability in lineup evaluations. Further investigations on how these relate to known  tests of cognitive skills, such as e.g. the paper folding tests.

### 2012-06-27 review/experiment-two.tex
revision

--- 2012-03-31 experiment-two.tex
@@
-% !TEX root = comparison.tex
-\subsection{Comparing Distribution Centers}
+% !TEX root = comparison-review.tex
+\subsection{Study II: Comparing Distribution Centers}
+\subsubsection{Study Setup}
+In a second experiment, we conducted a simulation study to investigate the power of four competing designs, as shown in Figure \ref{fig:expii}, in assessing a mean shift between distributions.
+
+\noindent The  hypotheses for each lineup are
+
+\begin{tabular}{rl}
+$H_o:$&{\em centers of the two groups are the same}, \\
+$H_a:$&{\em center of the blue group is shifted to the right}.\\[3pt]
+\end{tabular}
+%\vspace{-0.1in}
+%\vspace{-0.1in}
+
+As covariates for this experiment, we considered the {\it size of the shift}, $d$, between the means of the distributions, with $d=0.4, 0.6, 0.8.
+1.0,$ and $1.2$, as well as the sample size.
+Since we are dealing with two groups of points, we varied both the {\em size of the larger group}, $n_1 = 15, 45$, and 135, as well as the {\em relative size} of the second group ({\em ratio}), with $n_2 = 1/3, 2/3$ and $3/3$ of $n_1$.
+For each of these combinations, we created three replicate data sets with values from an exponential distribution with rate parameter $\lambda=1$ (for the first group) and rate parameter $\lambda_2=1/(d+1)$, which corresponds to a mean shift of $d$, for the second groups.
+Based on each data set we created four lineup charts, one for each of the competing designs.
+This resulted in 540 different lineup charts overall, which we sorted by $p$-value corresponding to the difference in means between the two groups in the simulated data.
+According to the $p$-value, we associated a `difficulty' level with lineups from 1 to 9 (easy to hard).
+Ten lineup plots were shown to each participant - one out of each of the difficulty levels and one additional reference chart with a particularly easy lineup  to allow for data quality checks.
+\begin{figure} [hbtp]
+   \centering
+   \includegraphics[width=0.45\linewidth]{hist-conc} 
+   \includegraphics[width=0.45 \linewidth]{boxplot-conc} \\
+   \includegraphics[width=0.45 \linewidth]{density-conc} 
+   \includegraphics[width=0.45 \linewidth]{dotplot-conc} 
+\vspace{-0.1in}
+   \caption{Overview of all four competing designs for displaying differences in distributions.}
+   \label{fig:expii}
+\end{figure}
+
+All of the designs were chosen based on common usage. The main difference between the designs is their varying level of data summarization.
+
+ From most to least summarized we consider these designs:
+
+\begin{tabular}{lp{0.33\textwidth}}\\[-5pt]
+\em boxplots: & five-point summary statistics (and a number of  outliers),\\
+\em density plots: & need -- mathematically -- all of the data points, but  their smoothness and continuity makes them appear visually quite simple,\\
+\end{tabular}
+\begin{tabular}{lp{0.33\textwidth}}
+ {\em histograms}: &  rendered by a frequency breakdown for a specified number of bins ($10 \times 2$, in the study),\\
+ {\em dotplots}: & all $n$ data points necessary for rendering.\\[3pt]
+\end{tabular}
+
+\noindent For the boxplot and dotplot, the vertical axis is free to be used for another variable, {\em group}, so the data can be split apart.
+For the histogram and density the vertical axis is used for count or frequency, and is not available for group information.
+Based on the assumption that charts are most effective \citep{Kosslyn:2006}, when they show enough information to solve the task at hand, but not too much to hamper the observer, we anticipated results to reflect the above order.
+
+\subsubsection{Results}
 
 Since each participant provides results from nine lineups, we can use a generalized linear mixed effects model of power and account for individuals' abilities by including a subject-specific random intercept.
-We included all of the above described covariates  in the model, both as main effects and as two-way interaction with the design to assess their impact on the power of each of the designs.
+All of the above described covariates are considered  in the model, both as main effects and as two-way interaction with the design to assess their impact on the power of each of the designs.
 Figure \ref{fig:power2} summarizes the modeling results as shown in table \ref{tbl:power2}.
@@
 Generally, an increase of the second group also is beneficial to power.
-Unlike all of the other designs,  dotplots actually lose power significantly as the second group increases.
-The reason behind this phenomenon might be visual complexity: dotplots are the only design that show individuals.
+Dotplots are an exception to this -- they actually lose power significantly as the second group increases.
+This might be due to visual complexity, as discussed above.
+%The reason behind this phenomenon might be visual complexity \citep{Kosslyn:2006}: dotplots are the only design that show individuals.
 With an increase in group size, visual complexity increases with it, which could hamper the observer in evaluating shifts.
@@
 \end{center}
+\vspace{-0.2in}
 \caption{\label{tbl:power2}Results from a Generalized Linear Mixed Effects Model of power of a  lineup given design and all two-way interactions with effect size $d$, group size $n_1$, and ratio to second to first group size.
@@
 
+Visually, density plots seem to 
 The two highly summarized designs benefit the most from an increase in sample size: both density charts and boxplots profit significantly more than histograms and dotplots.
@@
 \end{center}
+\vspace{-0.2in}
 \caption{\label{tbl:speed2}Results from a Linear Mixed Effects Model of (log) time to answer a lineup given design and all two-way interactions with effect size $d$, group size $n_1$, and ratio to second to first group size.
@@
 \end{table}
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+   \includegraphics[width=0.95\linewidth]{conf-exp2} 
+\vspace{-0.25in}
+   \caption{Barcharts of marginal distributions of self-reported confidence by design. The distributions display prominent differences.}
+   \label{fig:conf-margins}
+\end{figure}
 
@@
 While we tried to accommodate for over-plotting by making use of alpha-blending and jittering, this might not have been sufficient for the large sample size.
-
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-   \includegraphics[width=\linewidth]{conf-exp2} 
-   \caption{Barcharts of marginal distributions of self-reported confidence by design. The distributions display prominent differences.}
-   \label{fig:conf-margins}
-\end{figure}
 

### 2012-06-27 review/setup.tex
revision

% !TEX root = comparison.tex
\section{Experiments}~\label{setup}

Two recent data analyses motivated the setup of this experiment. The first one arises from an analysis of US flight traffic data, and a finding related to how wind direction affects the efficiency of an airport.  The second one arose during the review of a paper claiming that it was impossible to statistically test for differences in center between two samples where sample size was small and from a non-normal population, with application to studying toxic waste sites. Using lineups containing dotplots we found it was possible to detect a difference, and we were curious to see if other display types (density, histogram, boxplot) might compare with dotplots.

%We want two discuss two applications, in which we are making use of lineups to compare competing designs. in the second example we are interested in visualizing mean shifts between two distributions.

\subsection{Data Collection}
Data for both studies was collected using Amazon's Mechanical Turk (MTurk) service. 
The website for the second study is available at \url{http://www.public.iastate.edu/~mahbub/feedback_turk5/homepage.html}. Figure \ref{fig:screen} shows a screen shot of the website's layout for a dotplot lineup.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=\linewidth]{website} 
\vspace{-.2in}
   \caption{Screenshot of the website for the second study. }
   \label{fig:screen}
\end{figure}

Each participant is shown ten lineups and asked to identify the plot that is the most different from the other plots. In the second study the `difference' is further specified as `In which plot is the blue group the furthest to the right?". 
 In both studies, though, the question is placed prominently above each lineup as a reminder of what participants are asked to look for.  Additionally, participants are asked for a reason for their choice. They can select from a list of choices particular to the lineup design, why they chose this particular plot as well as how confident they are on a scale of one to five, where higher values indicate higher confidence. Personal information on age group, gender, and education is collected on a voluntary basis.  The amount of time it takes the individual to answer is also recorded.  This  allows us to assess how design choices affect decoding \cite{cleveland:1994} in terms of accuracy and speed.

 In order to avoid people `gaming the system' as found in earlier studies \cite{heer:2010, kosara:2010}, two of the ten lineups participants were shown were specially prepared as a baseline of performance -- if participants fail to answer some of these baseline lineups, we require at least two correct answers for the remaining plots, which is indicative of a better-than-guessing performance, as inclusion criterion for the data evaluation.
